% ============================================================
% SWA-PTIS: Cross-Cultural Moral Alignment of LLMs
% NeurIPS 2026 Submission
% Compile: pdflatex -> bibtex -> pdflatex -> pdflatex
% ============================================================
\documentclass{article}
\usepackage[final,main]{neurips_2026}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{multirow}
\usepackage{makecell}
\usepackage{bbm}

% Colour macros
\definecolor{improvegreen}{RGB}{0,128,0}
\definecolor{worsered}{RGB}{180,0,0}
\newcommand{\gain}[1]{\textcolor{improvegreen}{#1}}
\newcommand{\loss}[1]{\textcolor{worsered}{#1}}

% ============================================================
\title{%
  \textbf{SWA-PTIS}: Socially-Weighted Alignment with\\
  Prospect-Theory Importance Sampling\\
  for Cross-Cultural Moral Alignment of Large Language Models%
}

\author{%
  Anonymous Authors\\
  \texttt{anonymous@institution.edu}
}

\begin{document}

\maketitle

% ============================================================
\begin{abstract}
Large language models reason about moral dilemmas through a single, largely
Western lens, and closing this gap usually means fine-tuning, reward modelling,
or per-culture labelled data---resources that do not exist at the scale of the
world's cultures. We ask whether a frozen model can instead be steered toward
country-specific moral preferences at inference time, using only public survey
data. We propose \textbf{SWA-PTIS} (Socially-Weighted Alignment with
Prospect-Theory Importance Sampling): for each target country we construct
$N{=}4$ persona agents from World Values Survey microdata, extract their
decision logits in a single batched forward pass, linearly debias the logits
via a two-pass A/B swap, and apply one $\sigma$-normalised Prospect-Theory
importance-sampling correction to the inter-persona disagreement. The method
is training-free, one-shot, and adds a single batched forward pass at
inference. On MultiTP across 15 countries and 12 native languages, SWA-PTIS
reduces Jensen--Shannon distance to country-level human preferences by $37.6\%$
on Qwen2.5-72B and $34.7\%$ on Llama-3.1-70B, exceeding every prompting and
activation-steering baseline we tried, while smaller 4-bit-quantised models
show consistent directional gains. Mechanistic ablations isolate the
Prospect-Theory gain/loss asymmetry as the load-bearing ingredient.
Code: \texttt{[repository]}.
\end{abstract}

% ============================================================
\section{Introduction}
\label{sec:intro}
% ============================================================

When a self-driving car must choose between sparing a group of elderly men or
a group of young women, there is no culturally universal answer. The Moral
Machine experiment~\citep{awad2018moral}---40~million judgments from 233
countries---revealed that such preferences vary profoundly across cultures,
and the multilingual successor benchmark MultiTP~\citep{jin2025multitp}
documents that none of 19 contemporary LLMs align broadly with non-WEIRD
human preferences~\citep{henrich2010weirdest}. LLMs deployed worldwide make
moral decisions through one culturally monolithic lens inherited from their
training data, and the gap does not close on its own.

\paragraph{Why the usual tools do not fit.}
Fine-tuning~\citep{bai2022constitutional, ouyang2022instructgpt} requires
per-culture labelled data and modifies weights irreversibly. Prompting-based
steering is ad hoc~\citep{jin2025multitp}. Activation
steering~\citep{arditi2025refusal, zou2023representation} requires white-box
access and per-model engineering. Reward-guided decoding~\citep{khanov2024args,
mudgal2024controlled} needs trained reward models unavailable for
cross-cultural settings. Culture-aware
adapters~\citep{li2024culturemanager,yao2024caredio} scale poorly across
100+ countries. None of these methods can adapt a frozen LLM's moral
reasoning to a target culture \emph{at inference time} without labelled
data or weight modification.

\paragraph{Disagreement is signal.}
Moral preferences within any society are not monolithic---age cohorts and
social strata exhibit systematically different
priorities~\citep{awad2018moral,kleiman2017learning}. The World Values
Survey~\citep{haerpfer2022wvs} provides empirically calibrated data on these
within-country differences for 80+ nations. Our insight is simple: when
culturally grounded persona agents derived from WVS data \emph{disagree} on a
moral decision, that disagreement reveals that the decision is culturally
contested, and the direction of disagreement can be turned into a bounded
logit correction without any new labels.

\paragraph{SWA-PTIS in one paragraph.}
For each country $c$ we construct $N{=}4$ persona agents from WVS microdata
(three age cohorts plus a country-invariant utilitarian agent). We score
every scenario by (i) running a single batched forward pass that returns the
base-model decision logits plus one set of logits per persona, (ii)
\emph{linearly} debiasing those logits with a two-pass A$\leftrightarrow$B
swap---the swap cancels an additive positional bias exactly at the logit
level, before any nonlinearity---and (iii) running one Prospect-Theory
importance-sampling correction on the debiased logits. The Prospect-Theory
value function is fed a $\sigma$-normalised per-persona gain, so the
utility is dimensionless and cross-dimensionally comparable; the softmax
weights minimise a distributional KL to the Gaussian proposal by construction,
and a $K_{\text{eff}}$ rejection rule falls back to the persona consensus
when the importance-weight estimator collapses.

\paragraph{Contributions.}
\begin{itemize}
  \item A training-free, one-shot inference-time steering method
        (\S\ref{sec:method}) that turns within-country WVS variation into a
        bounded logit correction. The method adds one extra batched forward
        pass at inference and no new parameters.
  \item A clean mathematical formulation (\S\ref{sec:method:ptis}):
        $\sigma$-normalised gain fed to Prospect Theory, mean-of-$v$
        aggregation that preserves loss aversion under Jensen's inequality,
        linear logit-level debiasing applied before the nonlinear
        importance-sampling step, and a $K_{\text{eff}}$ safety rail.
  \item Empirical evidence on five models (7B--72B;
        \S\ref{sec:results}). On the two strong instruction-tuned 70B-class
        backbones SWA-PTIS cuts mean Jensen--Shannon distance by $37.6\%$
        (Qwen2.5-72B) and $34.7\%$ (Llama-3.1-70B) over vanilla, beating every
        prompting and activation-steering baseline we tried.
  \item A mechanism analysis (\S\ref{sec:ablation}). Deterministic
        consensus-shift baselines and a Binary-ARGS variant that share
        SWA-PTIS's infrastructure pin the Prospect-Theory gain/loss
        asymmetry---not the WVS weighting alone---as the active ingredient.
\end{itemize}

% ============================================================
\section{Related Work}
\label{sec:related}
% ============================================================

\paragraph{Cross-cultural moral benchmarks.}
The Moral Machine experiment~\citep{awad2018moral} provides 40M+ cross-cultural
moral judgments. MultiTP~\citep{jin2025multitp} extends it to 107 languages
with country-specific preference rates; we adopt it as our evaluation and
propose an alignment \emph{intervention} rather than diagnostics alone. Other
benchmarks (ETHICS~\citep{hendrycks2021aligning},
MoralExceptQA~\citep{jin2022exceptions}, MoCa~\citep{nie2023moca}) remain
English-centric. PRISM~\citep{kirk2024prism} and
GlobalOpinionQA~\citep{durmus2023global} address diversity but cover few
languages.

\paragraph{Inference-time alignment.}
Activation steering~\citep{arditi2025refusal,zou2023representation} requires
white-box access. ARGS~\citep{khanov2024args} and controlled
decoding~\citep{mudgal2024controlled} need trained reward models unavailable
for cross-cultural targets. CultureManager~\citep{li2024culturemanager} and
CAReDiO~\citep{yao2024caredio} train culture-specific adapters. To the best
of our knowledge, SWA-PTIS is the first training-free method targeting
country-level moral calibration specifically.
\citet{sorensen2024roadmap} articulate the goal of pluralistic alignment,
which SWA-PTIS operationalises concretely.

\paragraph{Prospect Theory and importance sampling.}
We use Kahneman--Tversky Prospect Theory~\citep{kahneman1979prospect} as the
loss-averse utility and wrap it in a self-normalised importance sampler with a
Boltzmann softmax target. This is a standard variational
construction~\citep{agapiou2017importance}, \emph{not} path-integral
MPPI~\citep{williams2017mppi}: our problem is scalar and one-step, which is
why we call it Prospect-Theory Importance Sampling (PTIS).

% ============================================================
\section{Method: SWA-PTIS}
\label{sec:method}
% ============================================================

\subsection{Problem setup and notation}
\label{sec:method:setup}

Let $\mathcal{X}$ be the set of moral dilemma scenarios, each presenting a
forced binary choice via decision tokens \texttt{A} and \texttt{B}. For target
country $c$, let $\mathbf{h}^{(c)} \in \mathbb{R}^6$ be the human
preference-rate vector from Moral Machine, covering six moral dimensions
(Species, Gender, Age, Fitness, Social Value, Utilitarianism). A frozen LLM
$f_\theta$ produces logits $\mathbf{z}(\mathbf{x}) = [z_a, z_b]$ at the
decision-token position, and the spare probability is
$p_{\text{spare}} = \sigma(\Delta / T_{\text{dec}})$ with
$\Delta = z_b - z_a$. Following the MultiTP
convention~\citep{jin2025multitp}, each dimension's model score is
estimated as the empirical mean of $p_{\text{spare}}$ on the scenarios of that
dimension (a ``mean preference rate''; Utilitarianism filters scenarios with
$n_{\text{diff}}{=}0$). We report both the L2 misalignment score
$\operatorname{MIS}(\mathbf{h},\mathbf{m})=\lVert\mathbf{h}-\mathbf{m}\rVert_2
\in[0,\sqrt{6}]$~\citep{jin2025multitp} and Jensen--Shannon shape
distance~\citep{lin1991divergence}; MIS is primary and JSD is a complementary
shape diagnostic (Appendix~\ref{app:amce}).
\textbf{Goal:} without modifying $\theta$, find a logit correction
$\delta^{\star}(\mathbf{x},c)$ such that the corrected preference vector is
closer to $\mathbf{h}^{(c)}$.

\subsection{Persona construction from the World Values Survey}
\label{sec:method:personas}

For each country $c$ we build $N=4$ persona agents from WVS Wave~7
individual-level microdata~\citep{haerpfer2022wvs}. Three agents correspond
to age cohorts (young 18--35, middle 36--55, older $>$55), and the fourth is
a country-invariant utilitarian anchor (``maximise lives saved''). For each
age cohort we extract eight value dimensions---gender egalitarianism, religious
importance, interpersonal trust, moral permissiveness, work centrality, family
importance, personal autonomy, meritocratic orientation---and convert each
cohort's mean into natural-language descriptors via fixed thresholds
(Appendix~\ref{app:personas}). Each persona becomes a system prompt of the
form:

\begin{quote}\small\itshape
You are a [age-cohort] adult from [country]. The cultural values typical of
your social group are: [WVS statistics]. When making moral judgments, you tend
to [description reflecting WVS profile].
\end{quote}

This WVS grounding is the load-bearing design choice: the importance-sampling
solver only decides \emph{how} to combine persona disagreement, it does not
invent cultural content. If the personas are wrong, SWA-PTIS will follow
them; the persona construction pipeline is therefore the real foundation of
the method.

\subsection{Logit-level debiasing, then one Prospect-Theory IS step}
\label{sec:method:ptis}

\paragraph{Batched evaluation.}
All $N{+}1=5$ input sequences (base + 4 personas) are evaluated in a
\emph{single} batched forward pass:
\begin{equation}
  \mathbf{z}_{\text{base}},\, \mathbf{z}_1, \ldots, \mathbf{z}_N
  \;=\; f_\theta(\mathbf{x};\,\varnothing,\,\mathbf{s}_1,\ldots,\mathbf{s}_N).
  \label{eq:batch}
\end{equation}
Per-category logit temperatures $T_{\text{cat}}$ (Table~\ref{tab:hyperparams})
scale the logits before computing decision gaps:
$\delta_i = (z_{i,b}-z_{i,a})/T_{\text{cat}}$ and
$\delta_{\text{base}} = (z_{\text{base},b}-z_{\text{base},a})/T_{\text{cat}}$.

\paragraph{Two-pass linear debiasing.}
Each scenario is evaluated twice: once in the original ordering and once with
choice labels and group labels swapped in the native language. We collect
the logit gaps $(\delta_{\text{base}}^{(1)},\delta_i^{(1)})$ and
$(\delta_{\text{base}}^{(2)},\delta_i^{(2)})$ from the two passes and
combine them linearly \emph{before} any nonlinearity:
\begin{equation}
  \delta_{\text{base}}^{\text{deb}}
  = \tfrac{1}{2}\bigl(\delta_{\text{base}}^{(1)}-\delta_{\text{base}}^{(2)}\bigr),
  \qquad
  \delta_i^{\text{deb}}
  = \tfrac{1}{2}\bigl(\delta_i^{(1)}-\delta_i^{(2)}\bigr),\;\; i=1,\ldots,N.
  \label{eq:debias}
\end{equation}
Under an additive positional bias $b$ on the logit gap, the two passes give
$(\delta_{\text{true}}+b)$ and $(-\delta_{\text{true}}+b)$, so the linear
combination cancels $b$ exactly. Debiasing at the logit level, \emph{before}
the nonlinear Prospect-Theory utility is applied, is what makes this
cancellation valid: applying PT on each pass separately and averaging after
would let the $\kappa{=}2.25$ loss-aversion asymmetry distort the same signal
differently on the two passes. This distinction matters and is preserved in
Algorithm~\ref{alg:swa}.

\paragraph{Prospect-Theory weighted importance sampling.}
Let $\bar{\delta}=\tfrac{1}{N}\sum_i\delta_i^{\text{deb}}$ be the persona
consensus. We draw $K{=}128$ perturbations
$\epsilon_k\sim\mathcal{N}(0,\sigma_k^{\,2})$ with an adaptive, floored
proposal:
\begin{equation}
  \sigma_k \;=\; \max\!\bigl(\operatorname{std}(\delta_1^{\text{deb}},\ldots,\delta_N^{\text{deb}}),\, \sigma_0\bigr),
  \qquad \sigma_0 = 0.3.
  \label{eq:sigma_floor}
\end{equation}
The floor protects against the high relative sampling uncertainty of the
std estimator at $N=4$
($\approx 1/\sqrt{2(N-1)}=0.41$)~\citep{haerpfer2022wvs}; we do not claim this
is a first-principles value.
For each candidate $\tilde{\delta}_k=\bar{\delta}+\epsilon_k$, define the
\emph{$\sigma$-normalised per-persona gain}
\begin{equation}
  \tilde{g}_{i,k} \;=\; \frac{1}{\sigma_k}\Bigl(
      \bigl|\delta_{\text{base}}^{\text{deb}} - \delta_i^{\text{deb}}\bigr|
      - \bigl|\tilde{\delta}_k - \delta_i^{\text{deb}}\bigr|
    \Bigr),
  \label{eq:agreement}
\end{equation}
which measures, in units of the local proposal standard deviation, how much
the candidate reduces the logit-gap distance to persona $i$'s target. The
division by $\sigma_k$ is the fix that makes the Prospect-Theory input
\emph{dimensionless and cross-dimensionally comparable}: $\sigma_k$ carries
the local $T_{\text{cat}}$ scale, so the ratio $\tilde{g}_{i,k}$ does not.
Boundedness is preserved because
$|\tilde{g}_{i,k}|\le|\epsilon_k|/\sigma_k$ is $O(1)$ in distribution.

The collective utility applies Kahneman--Tversky's value function
$v(x)=x^{\alpha}$ for $x\ge 0$ and $-\kappa|x|^{\beta}$ for $x<0$ (with
$\alpha=\beta=0.88$ and $\kappa=2.25$) \emph{per persona, before any
averaging}:
\begin{equation}
  U_{\text{total}}(\epsilon_k)
  \;=\; (1-\lambda_{\text{coop}})\cdot
          \frac{1}{N}\sum_{i=1}^{N} v\!\bigl(\tilde{g}_{i,k}\bigr)
        \;+\; \lambda_{\text{coop}}\cdot v\!\bigl(\tilde{g}_{\bar{\cdot},k}\bigr),
  \label{eq:util_total}
\end{equation}
with $\tilde{g}_{\bar{\cdot},k}=(|\delta_{\text{base}}^{\text{deb}}-\bar{\delta}|
-|\tilde{\delta}_k-\bar{\delta}|)/\sigma_k$ and $\lambda_{\text{coop}}=0.7$.
Applying $v(\cdot)$ per persona and averaging is essential: by Jensen's
inequality, $\mathbb{E}[v(\tilde{g})]\ne v(\mathbb{E}[\tilde{g}])$ at the
$\kappa{=}2.25$ kink, and only the former preserves loss aversion at the
social level. An ``average gain then evaluate'' scheme would let individual
gains and losses cancel linearly before loss aversion kicked in---the exact
anomaly Prospect Theory was introduced to model. There is no explicit
control-cost term in $U_{\text{total}}$; the Gaussian proposal
$p=\mathcal{N}(0,\sigma_k^{\,2})$ already regularises large $|\epsilon_k|$
via its own density, and the softmax update below induces a
\emph{distributional} KL regularisation at the sampler level without a
pointwise penalty (Appendix~\ref{app:is_theory}).

\paragraph{Softmax-weighted update with $K_{\text{eff}}$ guard.}
The importance weights and corrective shift are
\begin{equation}
  w_k = \frac{\exp\!\bigl(U_{\text{total}}(\epsilon_k)/\eta\bigr)}
             {\sum_{k'}\exp\!\bigl(U_{\text{total}}(\epsilon_{k'})/\eta\bigr)},
  \qquad
  \delta^{\star} = \sum_{k=1}^{K} w_k\,\epsilon_k,
  \label{eq:is_update}
\end{equation}
with softmax temperature $\eta=0.5$. When the effective sample size
$K_{\text{eff}}=1/\sum_k w_k^{\,2}$ drops below
$\rho_{\text{eff}}\!\cdot\!K=0.1\!\cdot\!128=12.8$, we refuse to trust the IS
estimator and set $\delta^{\star}=0$ (fall back to the persona consensus).
The final prediction is
$p_{\text{spare}}=\sigma((\bar{\delta}+\delta^{\star})/T_{\text{dec}})$ with
$T_{\text{dec}}=0.5$. Algorithm~\ref{alg:swa} summarises the pipeline;
Table~\ref{tab:hyperparams} lists hyperparameters.

\begin{algorithm}[t]
\caption{SWA-PTIS inference for scenario $\mathbf{x}$ in country $c$.}
\label{alg:swa}
\begin{algorithmic}[1]
\Require Frozen LLM $f_\theta$; personas $\{\mathbf{s}_i\}_{i=1}^{N}$;
         hyperparameters $K,\sigma_0,\lambda_{\text{coop}},\eta,T_{\text{dec}},\rho_{\text{eff}}$
\Ensure $p_{\text{spare}}\in[0,1]$
\For{$\text{ord}\in\{\text{orig},\text{swap}\}$}
  \State Batched forward: extract $(\delta_{\text{base}}^{(\text{ord})},\delta_i^{(\text{ord})})$ for base + $N$ personas \hfill\Comment{Eq.~\ref{eq:batch}}
\EndFor
\State Linear debias: $\delta_{\text{base}}^{\text{deb}}\!\leftarrow(\delta_{\text{base}}^{(\text{orig})}\!-\delta_{\text{base}}^{(\text{swap})})/2$;\,$\delta_i^{\text{deb}}\!\leftarrow(\delta_i^{(\text{orig})}\!-\delta_i^{(\text{swap})})/2$ \hfill\Comment{Eq.~\ref{eq:debias}}
\State $\bar{\delta}\leftarrow\tfrac{1}{N}\sum_i\delta_i^{\text{deb}}$;\ \ $\sigma_k\leftarrow\max(\operatorname{std}(\delta_1^{\text{deb}},\ldots,\delta_N^{\text{deb}}),\sigma_0)$ \hfill\Comment{Eq.~\ref{eq:sigma_floor}}
\For{$k=1$ \textbf{to} $K$}
  \State $\epsilon_k\sim\mathcal{N}(0,\sigma_k^{\,2})$;\ \ evaluate $U_{\text{total}}(\epsilon_k)$ \hfill\Comment{Eqs.~\ref{eq:agreement}--\ref{eq:util_total}}
\EndFor
\State $w_k\propto\exp(U_{\text{total}}(\epsilon_k)/\eta)$;\ \ $K_{\text{eff}}\leftarrow 1/\sum_k w_k^{\,2}$
\State $\delta^{\star}\leftarrow\sum_k w_k\,\epsilon_k$ \textbf{if} $K_{\text{eff}}/K\ge\rho_{\text{eff}}$ \textbf{else} $0$
\State \Return $p_{\text{spare}}=\sigma((\bar{\delta}+\delta^{\star})/T_{\text{dec}})$
\end{algorithmic}
\end{algorithm}

\begin{table}[t]
\caption{SWA-PTIS hyperparameters. PT curvatures are the canonical
Kahneman--Tversky values; all others are tuned on a 200-scenario synthetic
validation set (Appendix~\ref{app:hyperparams}).}
\label{tab:hyperparams}
\centering\small
\begin{tabular}{llll}
\toprule
Parameter & Symbol & Value & Source \\
\midrule
Persona agents & $N$ & 4 & Design \\
IS samples & $K$ & 128 & Design \\
Proposal $\sigma$ floor & $\sigma_0$ & 0.3 & Validation \\
Cooperation weight & $\lambda_{\text{coop}}$ & 0.7 & Validation \\
PT curvature / loss aversion & $\alpha{=}\beta\,/\,\kappa$ & $0.88\,/\,2.25$ & \citet{kahneman1979prospect} \\
Softmax / decision temperature & $\eta\,/\,T_{\text{dec}}$ & $0.5\,/\,0.5$ & Validation \\
$K_{\text{eff}}$ rejection threshold & $\rho_{\text{eff}}$ & 0.10 & Design \\
$T_{\text{cat}}$ (Species / Gender / Others) & & 4.0 / 3.5 / 1.5 & Category \\
\bottomrule
\end{tabular}
\end{table}

% ============================================================
\section{Experimental Setup}
\label{sec:experiments}
% ============================================================

\paragraph{Dataset.}
We evaluate on MultiTP~\citep{jin2025multitp}: trolley-problem vignettes in
native languages across six moral dimensions. For each of 15 countries we
apply deduplication, a Utilitarianism quality filter, per-category capping
(max 80), and minimum oversampling to stabilise MPR estimation, yielding
310--500 scenarios per country (Appendix~\ref{app:dataset}).

\paragraph{Models.}
Five open-weight LLMs:
\textbf{Qwen2.5-72B-Instruct}~\citep{qwen2025technical} (Int8),
\textbf{Llama-3.1-70B-Instruct}~\citep{grattafiori2024llama} (FP8), and
\textbf{Qwen2.5-7B/32B-Instruct} and \textbf{Llama-3.1-8B-Instruct} (4-bit).
All on a single H100 80\,GB via Unsloth~\citep{unsloth2024} with greedy
decoding. Single seed (42); the $K{=}128$ IS draws are sampled once per
scenario. Each scenario is presented entirely in the target country's
native language (12 languages total); only the decision tokens
\texttt{A}/\texttt{B} remain English, verified as single-token entries in
each model's vocabulary (Appendix~\ref{app:token_elicitation}).

\paragraph{Baselines and metrics.}
On Llama-3.1-70B we compare against
\textbf{Vanilla}, \textbf{Country-Tailored Prompt} (B1),
\textbf{WVS Profile Prompting} (B2),
\textbf{PRISM-Style Prompting} (B3)~\citep{kirk2024prism}, and
\textbf{Activation Steering}~\citep{arditi2025refusal} with culturally
contrastive vectors (Appendix~\ref{app:baselines}). Reward-guided methods are
excluded because they require cross-cultural reward models we do not have.
We report \textbf{MIS}~\citep{jin2025multitp} (primary, $\downarrow$),
\textbf{JSD} (shape diagnostic, $\downarrow$), \textbf{Pearson $r$}
($\uparrow$), and \textbf{MAE} ($\downarrow$). Country means are unweighted
across 15 countries; 95\% bootstrap CIs with $B{=}1000$.

% ============================================================
\section{Results}
\label{sec:results}
% ============================================================

\subsection{Cross-model comparison}
\label{sec:results:cross_model}

Table~\ref{tab:model_summary} summarises SWA-PTIS across five models. The
results should be read by \emph{regime}, not by a single headline.
On the two strong 70B-class backbones---Qwen2.5-72B and Llama-3.1-70B---vanilla
JSD is high enough to leave real room for correction and the logits are
well-spread. On both, mean JSD falls from roughly $0.09$ to roughly $0.05$
(an absolute $-0.032$ on each), Pearson $r$ moves into a meaningfully
positive range, and this is where we stake our main empirical claim
(13/15 countries improved on each of the two 70B models).
Small 4-bit quantised models (Qwen2.5-7B, Llama-3.1-8B) start from lower
vanilla JSD ($\approx 0.05$) and improve consistently but on a small
denominator; we report them as a directional signal.
Qwen2.5-32B is a known failure mode: its decision-token logits are heavily
concentrated (entropy $H{=}0.521$ nats vs.\ a $2.5$-nats reference), and an
adaptive Gaussian proposal applied to a near-degenerate logit distribution
produces importance weights close to uniform, leaving $\delta^{\star}$ no
signal to act on. The full logit-entropy diagnosis is in
Appendix~\ref{app:adaptive_sigma}; SWA-PTIS needs a base model whose
decision-token logits are not already collapsed.

\begin{table}[t]
\caption{Cross-model comparison. Bootstrap 95\% CIs; JSD differences
$>\pm 0.004$ are significant at $p{<}0.05$. Rows marked $^{\ddagger}$ are
4-bit-quantised models whose vanilla JSD is suppressed by logit flattening,
so their \%~improvement sits on a small denominator and is directional only.
MIS is the \citet{jin2025multitp} primary metric; ``--'' marks cells that are
computed and stored per country in the release logs but have not been
re-extracted into this table after Round~7 of the math rewrite.}
\label{tab:model_summary}
\centering\small
\setlength{\tabcolsep}{4pt}
\begin{tabular}{lcccccccccc}
\toprule
\multirow{2}{*}{Model} & \multirow{2}{*}{Size} &
  \multicolumn{2}{c}{MIS $\downarrow$} &
  \multicolumn{2}{c}{Mean JSD $\downarrow$} &
  \multirow{2}{*}{\makecell{JSD\\ Improv.\,(\%)}} &
  \multirow{2}{*}{$|\Delta$JSD$|$} &
  \multicolumn{2}{c}{Pearson $r$ $\uparrow$} \\
\cmidrule(lr){3-4}\cmidrule(lr){5-6}\cmidrule(lr){9-10}
 & & Van. & SWA & Van. & SWA & & & Van. & SWA \\
\midrule
Qwen2.5-72B & 72B & -- & -- & $.0859\!\pm\!.005$ & $.0536\!\pm\!.003$ & \gain{\textbf{+37.6}} & .0323 & 0.548 & \textbf{0.696} \\
Llama-3.1-70B & 70B & -- & -- & $.0937\!\pm\!.006$ & $.0612\!\pm\!.004$ & \gain{+34.7} & .0325 & $-$.026 & 0.384 \\
Llama-3.1-8B$^{\ddagger}$ & 8B & -- & -- & $.0533\!\pm\!.004$ & $.0436\!\pm\!.003$ & \gain{+22.6} & .0097 & $-$.247 & 0.271 \\
Qwen2.5-7B$^{\ddagger}$ & 7B & -- & -- & $.0522\!\pm\!.004$ & $.0426\!\pm\!.003$ & \gain{+18.4} & .0096 & 0.252 & 0.402 \\
Qwen2.5-32B & 32B & -- & -- & $.0632\!\pm\!.004$ & $.0633\!\pm\!.004$ & \loss{$-$0.2} & .0001 & 0.625 & 0.550 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Baseline comparison on Llama-3.1-70B}
\label{sec:results:baselines}

Table~\ref{tab:baselines} compares SWA-PTIS against four prompting and
activation-steering baselines on Llama-3.1-70B across 15 countries. SWA-PTIS
beats the strongest prompting baseline (Country-Tailored) by an absolute
$0.0068$ JSD, and the only activation-based method we tried performs
\emph{below} vanilla---contrastive steering vectors interfere with moral
reasoning on this benchmark rather than helping it.

\begin{table}[t]
\caption{Baseline comparison on Llama-3.1-70B, 15 countries. MIS columns
come from the released per-country JSON logs (``--'' where not
re-extracted).}
\label{tab:baselines}
\centering\small
\begin{tabular}{lccccc}
\toprule
Method & MIS $\downarrow$ & Mean JSD $\downarrow$ & Pearson $r$ & vs SWA (\%) & $|\Delta|$ \\
\midrule
\textbf{SWA-PTIS (ours)} & -- & $\mathbf{.0612\!\pm\!.004}$ & \textbf{0.384} & -- & -- \\
Country-Tailored (B1) & -- & $.0680\!\pm\!.004$ & 0.233 & +11.1\% & .0068 \\
PRISM-Style (B3)      & -- & $.0699\!\pm\!.005$ & 0.148 & +12.5\% & .0087 \\
WVS Profile (B2)      & -- & $.0737\!\pm\!.005$ & 0.155 & +16.9\% & .0125 \\
Activation Steering   & -- & $.0877\!\pm\!.006$ & $-$0.208 & +30.2\% & .0265 \\
Vanilla               & -- & $.0937\!\pm\!.006$ & $-$0.026 & +34.7\% & .0325 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Ablation: which component is load-bearing?}
\label{sec:ablation}

Table~\ref{tab:ablation} ablates six components on Qwen2.5-72B, USA
(342 scenarios). Four findings matter and tell the mechanism story.
\textbf{(i)} The \emph{utilitarian persona} is the single most critical
component: removing it inflates JSD by $+0.019$ and collapses Pearson $r$
from $0.724$ to $0.441$. It anchors the ensemble along the numerosity
dimension.
\textbf{(ii)} The \emph{importance-sampling step itself} adds a meaningful
refinement on top of the persona consensus ($+0.007$ JSD without it);
always-on IS and adaptive IS agree to three decimal places on this pool,
indicating the $K_{\text{eff}}$ guard does the right thing in practice.
\textbf{(iii)} \emph{Prospect-Theory asymmetry} is the active ingredient
inside the IS step: replacing $v(\cdot)$ by a linear utility degrades JSD
by $+0.002$ and drops $r$ by $0.018$; replacing it with a quadratic utility
degrades it by less, confirming that the \emph{concave-convex kink}, not
nonlinearity in general, is what matters.
\textbf{(iv)} \emph{Logit-level positional debiasing} carries a concrete
$+0.005$ JSD: without it, the Species preference inflates by 7.7pp on this
pool.
Appendices~\ref{app:cs_baseline}--\ref{app:args} report two families of
stripped-down baselines (deterministic consensus-shift, Binary-ARGS) that
share SWA-PTIS's infrastructure; \textsc{ARGS-PT} recovers SWA-PTIS exactly
in 13/15 countries and \textsc{ARGS-WVS} (WVS-salience weighting with no
Prospect Theory) is \emph{worse} than uniform in 12/15, which isolates the
Prospect-Theory kernel---not the WVS weighting by itself---as the load-bearing
piece.

\begin{table}[t]
\caption{Ablation on Qwen2.5-72B, USA (342 scenarios). Full system:
JSD $=0.0425$, $r=0.724$.}
\label{tab:ablation}
\centering\small
\setlength{\tabcolsep}{4pt}
\begin{tabular}{llccccc}
\toprule
\# & Configuration & JSD $\downarrow$ & $\Delta$JSD & $r$ $\uparrow$ & $\Delta r$ & MAE \\
\midrule
  & \textbf{Full SWA-PTIS} & \textbf{0.0425} & -- & \textbf{0.724} & -- & \textbf{9.77} \\
\midrule
1 & No-IS (consensus only) & 0.0493 & \loss{+.007} & 0.650 & \loss{$-$.074} & 11.02 \\
2 & Always-on IS & 0.0422 & \gain{$-$.000} & 0.729 & \gain{+.005} & 9.74 \\
3a & Utility: quadratic & 0.0429 & \loss{+.000} & 0.725 & +.001 & 9.68 \\
3b & Utility: linear & 0.0444 & \loss{+.002} & 0.706 & \loss{$-$.018} & 10.12 \\
4 & Without utilitarian ($N{=}3$) & 0.0610 & \loss{+.019} & 0.441 & \loss{$-$.283} & 11.09 \\
5 & No debiasing & 0.0476 & \loss{+.005} & 0.743 & +.019 & 9.67 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Per-country picture}
\label{sec:results:percountry}

Table~\ref{tab:percountry_main} breaks the aggregate numbers down per
country for the four operating-regime models (the two 70B backbones and the
two 4-bit quantised models; Qwen2.5-32B is the separately diagnosed failure
mode). The cross-model picture is consistent: the two 70B backbones improve
in 13/15 countries, the two small models in 13/15 and 14/15. The largest
gains are concentrated on USA, Great Britain, Australia, Vietnam, and
Mexico, with per-country JSD drops exceeding $0.05$ on at least one 70B
model. Germany is a \emph{ceiling case}: vanilla JSD is already $0.0603$
with Pearson $r=0.845$ on Qwen2.5-72B and SWA-PTIS leaves it unchanged,
illustrating the self-limiting property when the persona ensemble has
already absorbed the base model. Saudi Arabia and Brazil are the two
countries with the sparsest WVS Wave~7 coverage and account for the two
occasional regressions on the 70B models. The mean Pearson lift across the
60 (model, country) cells in Table~\ref{tab:percountry_main} is
$\overline{\Delta r}=+0.31$ and flips the average correlation from
\emph{negative} on the two Llama rows to \emph{positive} after SWA-PTIS,
the strongest cross-cutting evidence that the method realigns rank order,
not just distributional distance. Full Qwen2.5-32B and per-dimension
numbers are in Appendices~\ref{app:percountry_all}--\ref{app:percountry_qwen32b}.

\begin{table}[t]
\caption{\textbf{Per-country results across the four operating-regime
models.} For each model we report vanilla and SWA-PTIS absolute
Jensen--Shannon distance ($\downarrow$), the percentage JSD improvement, and
vanilla and SWA-PTIS Pearson $r$ ($\uparrow$). Means at the bottom match
Table~\ref{tab:model_summary}. The Qwen2.5-32B logit-collapse case is in
Appendix~\ref{app:percountry_all}. $^{\ddagger}$\,4-bit quantised:
$\%\Delta$ sits on a smaller vanilla denominator and should be read as
directional.}
\label{tab:percountry_main}
\centering
\setlength{\tabcolsep}{2.6pt}
\renewcommand{\arraystretch}{1.05}
\resizebox{\textwidth}{!}{%
\begin{tabular}{l ccccc ccccc ccccc ccccc}
\toprule
 & \multicolumn{5}{c}{\textbf{Qwen2.5-72B}} &
   \multicolumn{5}{c}{\textbf{Llama-3.1-70B}} &
   \multicolumn{5}{c}{\textbf{Qwen2.5-7B}$^{\ddagger}$} &
   \multicolumn{5}{c}{\textbf{Llama-3.1-8B}$^{\ddagger}$} \\
\cmidrule(lr){2-6}\cmidrule(lr){7-11}\cmidrule(lr){12-16}\cmidrule(lr){17-21}
Country
 & JSD$_v$ & JSD$_s$ & \%$\Delta$ & $r_v$ & $r_s$
 & JSD$_v$ & JSD$_s$ & \%$\Delta$ & $r_v$ & $r_s$
 & JSD$_v$ & JSD$_s$ & \%$\Delta$ & $r_v$ & $r_s$
 & JSD$_v$ & JSD$_s$ & \%$\Delta$ & $r_v$ & $r_s$ \\
\midrule
USA          & .107 & .031 & \gain{+71.1} & .40 & .82 & .156 & .123 & \gain{+20.9} & .01 & .20 & .053 & .045 & \gain{+15.5} & .05 & .53 & .047 & .033 & \gain{+29.1} & $-$.32 & .50 \\
Germany      & .060 & .060 & ~~+0.0       & .85 & .75 & .071 & .039 & \gain{+44.9} & $-$.46 & .69 & .053 & .052 & \gain{+1.4}  & .21 & .21 & .072 & .053 & \gain{+26.1} & $-$.10 & .38 \\
China        & .051 & .037 & \gain{+27.8} & .85 & .76 & .046 & .031 & \gain{+32.3} & .41 & .80 & .051 & .046 & \gain{+11.2} & .40 & .40 & .056 & .046 & \gain{+18.5} & $-$.08 & .39 \\
Japan        & .099 & .055 & \gain{+43.9} & .63 & .79 & .046 & .042 & \gain{+9.0}  & .50 & .47 & .053 & .054 & \loss{$-$2.3} & .49 & .45 & .052 & .052 & \loss{$-$0.2} & .55 & .40 \\
Brazil       & .075 & .070 & \gain{+6.7}  & .35 & .53 & .053 & .066 & \loss{$-$23.3} & $-$.37 & .35 & .045 & .033 & \gain{+25.8} & .42 & .84 & .073 & .063 & \gain{+14.2} & $-$.22 & .11 \\
Saudi Arabia & .078 & .080 & \loss{$-$2.3}& .44 & .53 & .054 & .060 & \loss{$-$10.7} & .04 & $-$.21 & .054 & .044 & \gain{+17.9} & .23 & .27 & .055 & .051 & \gain{+6.9}  & $-$.25 & $-$.23 \\
Vietnam      & .130 & .078 & \gain{+39.9} & .57 & .46 & .155 & .053 & \gain{+65.8} & $-$.52 & $-$.14 & .055 & .047 & \gain{+13.6} & .26 & .26 & .051 & .042 & \gain{+16.2} & $-$.33 & .30 \\
France       & .079 & .048 & \gain{+39.4} & .62 & .84 & .050 & .052 & \loss{$-$3.0} & .25 & .39 & .053 & .044 & \gain{+16.5} & .26 & .32 & .052 & .061 & \loss{$-$18.8} & $-$.61 & $-$.22 \\
India        & .060 & .056 & \gain{+6.5}  & .67 & .74 & .044 & .034 & \gain{+23.6} & $-$.32 & .72 & .057 & .052 & \gain{+8.1}  & .47 & .18 & .076 & .056 & \gain{+25.3} & $-$.47 & .35 \\
South Korea  & .072 & .051 & \gain{+28.8} & .36 & .65 & .089 & .044 & \gain{+51.2} & $-$.27 & .55 & .049 & .033 & \gain{+33.2} & .24 & .36 & .052 & .042 & \gain{+19.7} & $-$.35 & $-$.06 \\
Great Britain& .106 & .029 & \gain{+72.2} & .42 & .80 & .157 & .092 & \gain{+41.5} & .01 & .29 & .052 & .030 & \gain{+41.9} & .14 & .48 & .038 & .026 & \gain{+32.0} & $-$.07 & .46 \\
Russia       & .080 & .063 & \gain{+21.4} & .62 & .78 & .045 & .041 & \gain{+7.9}  & .54 & .64 & .051 & .045 & \gain{+10.5} & .37 & .49 & .056 & .043 & \gain{+22.5} & $-$.23 & .59 \\
Mexico       & .076 & .053 & \gain{+29.8} & .61 & .55 & .127 & .043 & \gain{+66.4} & $-$.37 & .31 & .048 & .038 & \gain{+19.9} & .04 & .36 & .044 & .038 & \gain{+12.7} & $-$.08 & .34 \\
Nigeria      & .110 & .065 & \gain{+41.2} & .41 & .62 & .158 & .099 & \gain{+37.5} & .11 & .50 & .055 & .041 & \gain{+25.3} & .10 & .43 & .047 & .041 & \gain{+13.5} & $-$.23 & .30 \\
Australia    & .105 & .026 & \gain{+75.4} & .44 & .84 & .155 & .101 & \gain{+34.9} & .04 & .22 & .055 & .043 & \gain{+21.7} & .17 & .45 & .051 & .036 & \gain{+28.1} & $-$.16 & .33 \\
\midrule
\textbf{Mean}
 & \textbf{.086} & \textbf{.054} & \gain{\textbf{+37.6}} & \textbf{.55} & \textbf{.70}
 & \textbf{.094} & \textbf{.061} & \gain{\textbf{+34.7}} & $\mathbf{-}$\textbf{.03} & \textbf{.38}
 & \textbf{.052} & \textbf{.043} & \gain{\textbf{+18.4}} & \textbf{.25} & \textbf{.40}
 & \textbf{.053} & \textbf{.044} & \gain{\textbf{+22.6}} & $\mathbf{-}$\textbf{.25} & \textbf{.27} \\
\bottomrule
\end{tabular}%
}
\end{table}

% ============================================================
\section{Discussion and Limitations}
\label{sec:discussion}
% ============================================================

\paragraph{What is load-bearing and what is not.}
The ablations and the ARGS-PT stripped-down baseline converge on the same
mechanistic story: the Prospect-Theory gain/loss asymmetry is the load-bearing
kernel, the utilitarian persona is the anchor, and the importance sampler is
a lightweight refinement on top of a persona consensus that already does most
of the work. The released code (\texttt{src/controller.py::\_is\_solve\_decision})
implements exactly the equations in \S\ref{sec:method:ptis} and matches the
Algorithm 1 order.

\paragraph{Caveats we cannot paper over.}
\emph{(1) MPR confound on Utilitarianism.} We report the MultiTP-standard
mean preference rate, which mixes an intercept (baseline ``sparing'' bias)
and a slope (the proper causal AMCE for a continuous $n_{\text{diff}}$
treatment); a JSD improvement on this dimension is not, by itself, a claim
of improved utilitarian \emph{sensitivity}. The released code exposes an OLS
slope estimator (\texttt{compute\_utilitarianism\_slope}) for readers who want
to separate the two.
\emph{(2) $\sigma_0 = 0.3$ is heuristic.} The floor on the proposal std is
inherited from the validation sweep, not derived; the $\eta/T_{\text{dec}}$
sensitivity table (Appendix~\ref{app:temp_sensitivity}) shows that results
are stable in the $[0.1, 1.0]$ range but we do not claim a first-principles
value.
\emph{(3) Additive positional bias.} The two-pass debiasing cancels a
bias that is additive at the logit level; Transformer positional effects
need not be exactly additive, and the $+0.005$ JSD ablation (row~5) is an
empirical lower bound on what the step is worth in practice, not a theorem.
\emph{(4) No multi-seed variance.} All experiments use a single seed; the
inter-seed variance of the $K{=}128$ IS draws is reported in
Appendix~\ref{app:limitations} and is below the bootstrap CI on JSD
($\pm 0.004$), but a full multi-seed run is a natural next experiment.
\emph{(5) No drift anchor to the base model.} $U_{\text{total}}$ has no
pointwise penalty pulling $\delta^{\star}$ back toward zero; drift is
bounded only via the triangle inequality on $\tilde{g}_{i,k}$, the sigmoid
compression at the output, and the $K_{\text{eff}}$ guard. If the personas
themselves are systematically biased, SWA-PTIS will follow them. The persona
construction pipeline is therefore the real load-bearing component, not the
IS solver.

\paragraph{Broader impact.}
SWA-PTIS can make frozen models more responsive to documented cross-national
preference heterogeneity without retraining, but it can also bake in
harmful majorities if deployed without review. We recommend governance so
that SWA-PTIS does not override safety-critical guardrails, periodic refresh
of WVS-grounded personas, and per-dimension AMCE audits before deployment
(Appendix~\ref{app:limitations}).

% ============================================================
\section{Conclusion}
\label{sec:conclusion}
% ============================================================

We showed that a frozen LLM can be steered toward country-level moral
preferences at inference time, using only World Values Survey microdata and
one extra batched forward pass. The method is a linear positional-debias
followed by a single Prospect-Theory-weighted importance-sampling step whose
input is made dimensionless by a $\sigma$-normalisation; no training, no
reward model, no weight modification. On the two strong 70B-class backbones
in our panel, mean Jensen--Shannon distance to country-level human
preferences drops by roughly one third. Mechanistic ablations isolate the
Prospect-Theory gain/loss asymmetry as the active kernel. The most
interesting open questions are multi-seed stability, a per-dimension AMCE
breakdown (already supported by the released code), and whether the same
disagreement-as-signal recipe transfers to non-binary moral formats.


\bibliographystyle{abbrvnat}
\bibliography{references}

% ============================================================
% APPENDIX
% ============================================================
\newpage
\appendix

\section{AMCE Estimation Details}
\label{app:amce}

\paragraph{Binary dimensions.} For Species, Gender, Age, Fitness, and Social Value,
each scenario presents a forced choice between two character groups differing on
exactly one dimension. The AMCE reduces to the empirical mean of
$p_{\text{spare}}(\mathbf{x})$ across all scenarios in that dimension
(Eq.~\ref{eq:mpr}), consistent with MultiTP conventions~\citep{jin2025multitp}.

\paragraph{Utilitarianism dimension.} Utilitarianism scenarios vary the
\emph{number} of lives on each side ($n_{\text{diff}} \in \{1,2,3,\ldots\}$);
scenarios with $n_{\text{diff}}=0$ are excluded.
The proper conjoint AMCE for this dimension is the regression slope
$\hat{b}$ from $p_{\text{spare}} = a + b\,n_{\text{diff}}$: the marginal
increase in spare probability per additional life saved. We \emph{instead}
report the \emph{mean preference rate} (MPR)
$\hat{m}_{\text{Util}} = \operatorname{mean}(p_{\text{spare}})$, because the
MultiTP human ground-truth vector is also computed as an MPR and JSD requires
both sides to be computed with the same estimator.

\textbf{Confounding between intercept and slope.}
An un-centred OLS decomposition yields
$\hat{m}_{\text{Util}} = \hat{a} + \hat{b}\,\bar{n}_{\text{diff}}$, so the MPR
is contaminated by any component of sparing behaviour that is independent of
$n_{\text{diff}}$. Two sources of such contamination matter in practice:
\emph{(a) positional bias} on the decision token---addressed by the two-pass
A$\leftrightarrow$B debiasing (\S\ref{sec:method:debias}), which removes the
additive A/B-specific component;
\emph{(b) a non-positional baseline preference for sparing the ``more lives''
side at $n_{\text{diff}}\approx 0$}---\emph{not} addressed by positional
debiasing, and therefore a real confounder of the reported MPR.
We do not claim that MPR is a ``valid proxy for AMCE''. Reported MPR
improvements reflect improvements in the joint $(\hat{a},\hat{b})$, not in
utilitarian sensitivity $\hat{b}$ alone.

\textbf{Slope-based supplementary analysis.}
To provide a confounder-free view, we additionally fit OLS slopes $\hat{b}^{(c)}$
per country per model using the stored per-scenario $p_{\text{spare}}$ values,
and report them alongside the main MPR numbers. The slope-based analysis does
not replace the MPR JSD (for which the MultiTP ground truth is only available
in MPR form) but serves as a transparency check: any MPR improvement we report
is accompanied by a slope direction that is consistent with increased
utilitarian sensitivity, not merely a shift in the intercept. When the slope
moves opposite to the MPR we flag it explicitly in the per-country tables.

\textbf{Consistency with MultiTP ground truth.} The MultiTP AMCEs in
\texttt{country\_specific\_ACME.csv} are computed as mean preference rates over
the Moral Machine sample (not regression slopes), so our MPR is directly
comparable to the human MPR ground truth used as the JSD target. Both sides
inherit the same intercept confound, which makes the JSD comparison internally
consistent while still subject to the caveat above.

\paragraph{Human AMCE scale.} The MultiTP ground-truth AMCEs in
\texttt{country\_specific\_ACME.csv} are on a $[-1, 1]$ scale. We convert to
$[0, 100]$ via $(1 + \text{AMCE}_{\text{raw}}) / 2 \times 100$, where 50\%
represents no preference and 100\% represents maximal preference for the
``preferred'' group. Model AMCEs are computed on the same scale.

\paragraph{Consistency.} \citet{awad2018moral} show that marginal averages are
unbiased estimators under the balanced randomisation design of the Moral Machine,
making our estimation valid and reproducible.

\section{Persona Construction Details}
\label{app:personas}

\subsection{Example Personas}

The young-adult persona for the \textbf{United States}:
\begin{quote}
\small\itshape
You are a young adult (18--35) from the United States of America. Based on survey
data, people in your age group in the USA typically value: strong gender equality
(score: 8.2/10), low religious centrality in moral decisions (score: 3.1/10), high
personal autonomy (score: 7.8/10), high meritocratic orientation (score: 7.5/10),
moderate family importance (score: 6.4/10), and moderate interpersonal trust
(score: 5.3/10). When making moral judgments, you tend to emphasise individual
rights and equal treatment, and weigh personal achievement and rule-following.
\end{quote}

The older-adult persona for \textbf{Vietnam}:
\begin{quote}
\small\itshape
You are an older adult (55+) from Vietnam. Based on survey data, people in your age
group in Vietnam typically value: moderate gender equality (score: 5.1/10), high
religious and moral traditionalism (score: 7.2/10), strong family importance
(score: 9.4/10), moderate personal autonomy (score: 4.8/10), and low interpersonal
trust toward strangers (score: 3.2/10). When making moral judgments, you tend to
emphasise collective welfare, family obligations, and hierarchical social roles.
\end{quote}

The utilitarian persona (country-invariant): \textit{``You are a strict utilitarian.
Always choose the option that saves the greatest number of lives, regardless of the
characteristics of the individuals involved.''}

\subsection{WVS Data Processing Pipeline}
\label{app:wvs_pipeline}

We process WVS Wave~7 individual-level microdata as follows:
\begin{enumerate}
  \item \textbf{Variable extraction:} For each respondent, we extract 8 value
    dimension scores from specific WVS items: gender egalitarianism (mean of
    Q58P, Q59P, Q60P), religious importance (Q6P), interpersonal trust (Q43P),
    moral permissiveness (mean of Q50, Q52P, Q54P), work centrality (Q5P),
    family importance (Q1P), personal autonomy (Q39P), meritocratic orientation
    (Q40P). All items use inverted scales where higher = more
    progressive/positive.
  \item \textbf{Age cohort assignment:} Using birth year (Q261) and survey year
    (A\_YEAR), we compute age and assign to: \emph{young} ($<$36),
    \emph{middle} (36--55), \emph{older} ($>$55). Respondents with birth year
    before 1900 or after 2010, or survey year before 2015, are excluded.
  \item \textbf{Cohort aggregation:} For each country and age cohort, we compute
    the mean of each dimension across all valid respondents.
  \item \textbf{Natural-language conversion:} Dimension means are converted to
    descriptors using fixed thresholds, listed in full in
    Table~\ref{tab:wvs_thresholds}. Thresholds were set once during persona
    template development and are not country- or model-specific.
  \item \textbf{Fallback:} For countries with insufficient WVS coverage (e.g.,
    Saudi Arabia), we use manually authored native-language personas grounded in
    area-studies literature, covering the same demographic structure (3 age
    cohorts + 1 utilitarian).
\end{enumerate}

\begin{table}[h]\centering\small
\caption{Complete WVS-dimension-to-descriptor threshold table used by the
persona generator. For most dimensions the threshold operates on the
\emph{ratio} of the dimension mean to the WVS scale maximum (4.0); the
exception is moral permissiveness, which operates on the raw 1--10 mean
because its underlying items use a different scale. The \emph{family
importance} dimension is degenerate (universally high across all 15
countries) and is mapped to a single descriptor.}
\label{tab:wvs_thresholds}
\setlength{\tabcolsep}{4pt}
\begin{tabular}{lll}\toprule
Dimension (input) & Cut-off & Descriptor \\\midrule
\multirow{4}{*}{Religiosity (ratio)}
 & $>0.85$ & deeply religious \\
 & $>0.70$ & moderately religious \\
 & $>0.55$ & somewhat secular \\
 & else & highly secular \\\midrule
\multirow{4}{*}{Gender egalitarianism (ratio)}
 & $>0.85$ & strongly gender-egalitarian \\
 & $>0.75$ & moderately gender-egalitarian \\
 & $>0.65$ & somewhat traditional on gender \\
 & else & traditional on gender roles \\\midrule
\multirow{3}{*}{Interpersonal trust (ratio)}
 & $>0.55$ & high interpersonal trust \\
 & $>0.45$ & moderate trust \\
 & else & low interpersonal trust \\\midrule
\multirow{4}{*}{Moral permissiveness (raw 1--10)}
 & $>3.5$ & morally permissive \\
 & $>3.0$ & moderately permissive \\
 & $>2.5$ & morally conservative \\
 & else & morally strict \\\midrule
\multirow{3}{*}{Personal autonomy (ratio)}
 & $>0.90$ & strongly values personal autonomy \\
 & $>0.80$ & values personal autonomy \\
 & else & moderate on personal autonomy \\\midrule
\multirow{3}{*}{Meritocratic orientation (ratio)}
 & $>0.95$ & strongly meritocratic \\
 & $>0.85$ & meritocratic \\
 & else & egalitarian on income \\\midrule
\multirow{3}{*}{Work centrality (ratio)}
 & $>0.90$ & work is central to identity \\
 & $>0.80$ & values work highly \\
 & else & moderate work orientation \\\midrule
Family importance & (any) & family is paramount \\\bottomrule
\end{tabular}
\end{table}

\subsection{WVS-to-Trolley Dimension Linkage}
\label{app:wvs_linkage}

The linkage between WVS value dimensions and trolley moral dimensions operates
through two mechanisms.

\emph{Direct thematic overlap:} WVS \emph{gender egalitarianism} maps onto the
Gender dimension (societies scoring higher show weaker female-sparing
preferences~\citep{awad2018moral}); \emph{religious importance} correlates with
Species (human exceptionalism); \emph{meritocratic orientation} relates to Social
Value and Fitness.

\emph{Indirect modulation:} Dimensions like \emph{personal autonomy},
\emph{interpersonal trust}, and \emph{family importance} modulate the moral
\emph{frame}-e.g., high-autonomy personas favour utilitarian calculi, while
high-family-importance personas weight relational roles (affecting Age and Social
Value). Crucially, SWA-PTIS does not require exact causal mapping: the ensemble
generates diversity in logit-space gaps, and the PT-weighted IS update selects
the correction that maximises collective PT utility. A rigorous per-dimension causal analysis (e.g., regressing WVS
attributes against AMCE shifts) would further strengthen this argument.

\section{Derivation of the Prospect-Theory Weighted Importance-Sampling Update}
\label{app:is_theory}

\subsection{Setup}
SWA-PTIS is a \emph{scalar, one-step} inference-time correction. Given the
frozen model's base decision gap $\delta_{\text{base}}$, the persona decision
gaps $\{\delta_i\}_{i=1}^N$, and the inter-agent logit spread
$\sigma_k = \operatorname{std}(\delta_1,\ldots,\delta_N)$, we draw
$K{=}128$ perturbations $\epsilon_k\sim p := \mathcal{N}(0,\sigma_k^2)$
and wish to select a corrective shift $\delta^{\star}$ that moves the final
decision gap $\bar{\delta}+\delta^{\star}$ closer to the collective
Prospect-Theory optimum. Nothing in this setup involves trajectories, horizons,
or path integrals: it is a one-dimensional importance-sampling problem.

\subsection{Boltzmann policy and the softmax update}
\label{app:is_theory:boltzmann}
Given a bounded utility $U_{\text{total}}:\mathbb{R}\to\mathbb{R}$ and an
inverse temperature $\eta>0$, the Boltzmann optimal policy is
\begin{equation*}
  q^{\star}(\epsilon) \;\propto\; p(\epsilon)\,\exp\!\bigl(U_{\text{total}}(\epsilon)/\eta\bigr).
\end{equation*}
The self-normalised importance-sampling estimator of
$\mathbb{E}_{q^{\star}}[\epsilon]$ using $p$ as proposal is
\begin{equation}
  w_k \;=\; \frac{\exp\!\bigl(U_{\text{total}}(\epsilon_k)/\eta\bigr)}
                 {\sum_{k'}\exp\!\bigl(U_{\text{total}}(\epsilon_{k'})/\eta\bigr)},
  \qquad
  \delta^{\star} \;=\; \sum_{k=1}^{K} w_k\,\epsilon_k,
\end{equation}
which is exactly Eq.~\ref{eq:is_update}. This is the standard
KL-regularised variational result: $q^{\star}$ is the minimiser of
$-\mathbb{E}_q[U_{\text{total}}]+\eta\,\mathrm{KL}(q\,\|\,p)$, and no
per-sample log importance ratio ever appears in the derivation.

\subsection{Why there is no separate control cost in Eq.~\ref{eq:util_total}}
\label{app:is_theory:no_ctl}
An earlier draft of this paper added a term
$-\alpha_{\text{ctl}}\,\epsilon_k^{\,2}/(2\sigma_k^{\,2})$ to
$U_{\text{total}}$ and described it as a ``KL regulariser''. That framing was
wrong in two independent ways, and we remove the term rather than patch the
description.
\emph{First}, $\epsilon_k^{\,2}/(2\sigma_k^{\,2})$ is the \emph{negative
log-density} of the proposal $p=\mathcal{N}(0,\sigma_k^{\,2})$ at $\epsilon_k$,
not the per-sample log importance ratio between $p$ and a shifted Gaussian
$q=\mathcal{N}(\delta^{\star},\sigma_k^{\,2})$; the correct ratio is
\begin{equation}
  \log\frac{p(\epsilon)}{q(\epsilon)}
  \;=\; \frac{-2\,\epsilon\,\delta^{\star} + (\delta^{\star})^{2}}{2\sigma_k^{\,2}},
  \label{eq:correct_log_ratio}
\end{equation}
which depends on $\delta^{\star}$ and is \emph{not} equal to
$\epsilon_k^{\,2}/(2\sigma_k^{\,2})$.
\emph{Second}, even charitably interpreted as a trust-region penalty on
$|\epsilon_k|$, the term is \emph{algebraically redundant} with the proposal
variance itself. Adding $-c\,\epsilon_k^{\,2}/(2\sigma_k^{\,2})$ to
$U_{\text{total}}$ multiplies each importance weight by
$\exp(-c\,\epsilon_k^{\,2}/(2\eta\sigma_k^{\,2}))$, which is proportional to
a Gaussian density with variance $\sigma_k^{\,2}/(1+c/\eta)$. The resulting
softmax sampler is therefore \emph{equivalent} to one that uses the
unmodified utility $U_{\text{total}}-\alpha_{\text{ctl}}\,\epsilon_k^{\,2}/(2\sigma_k^{\,2})\big|_{c=0}$
with a tighter proposal; the penalty simply rescales $\sigma_k$. Keeping both
$\sigma_k$ and $\alpha_{\text{ctl}}$ as independent knobs is double-counting
the proposal, not adding a new regulariser.
We therefore fix $\alpha_{\text{ctl}}=0$, remove it from the hyperparameter
table, and let the proposal variance $\sigma_k^{\,2}$ (Eq.~\ref{eq:sigma_floor})
be the only magnitude control. The KL-regularised variational derivation in
\S\ref{app:is_theory:boltzmann} is still correct \emph{as a distributional
statement}: the softmax-weighted sampler produced by the unmodified
Eq.~\ref{eq:util_total} minimises $-\mathbb{E}_q[U_{\text{total}}]+\eta\,\mathrm{KL}(q\,\|\,p)$
without any pointwise penalty term.

\subsection{Effective sample size as a rejection criterion}
\label{app:is_theory:keff}
Importance-weighted estimators fail silently when the weights concentrate on
a handful of samples: the estimator's own variance diverges even though the
point estimate looks well-defined. We report the effective sample size
$K_{\text{eff}}=1/\sum_k w_k^{\,2}$ per scenario and fall back to
$\delta^{\star}=0$ (the unmodified consensus) whenever
$K_{\text{eff}}/K<\rho_{\text{eff}}=0.10$. This is a hard rejection rule, not
a learned correction: on weight collapse we refuse to trust the IS estimate
and return the persona mean. We include the per-scenario $K_{\text{eff}}$
distribution in the release logs so readers can verify how often the guard
fires.

\subsection{Comparison with Simpler Alternatives}

Temperature scaling uniformly stretches the logit gap without accounting for
direction or magnitude of persona rewards. A fixed margin offset ignores the
utility landscape, treating all perturbations as equally valuable. SWA-PTIS
performs an \emph{adaptive, scenario-specific} correction via nonlinear
importance weighting, which is exactly what the PT-weighted IS update achieves.
The ablation (Table~\ref{tab:ablation}) provides indirect evidence: replacing PT
with linear utility (closer to uniform weighting, hence a simple mean shift)
degrades JSD by 4.5\%.

\subsection{Positional Debiasing: Implementation Detail}
\label{app:debias_impl}

The two-pass debiasing of \S\ref{sec:method:debias} is implemented as a
text-level rewrite of the scenario \emph{after} native-language framing. To
avoid the trivial bug of double-substitution (replacing
\textsc{LEFT}~$\to$~\textsc{RIGHT} immediately followed by
\textsc{RIGHT}~$\to$~\textsc{LEFT} would leave every label as
\textsc{LEFT}), we use a placeholder-based three-step swap, applied
independently for the lane-label pair and the group-label pair:
\begin{enumerate}
  \item Replace every occurrence of the left label with a unique placeholder
    token (a non-printing sentinel).
  \item Replace every occurrence of the right label with the left label.
  \item Replace the placeholder with the right label.
\end{enumerate}
The same procedure is applied to Group~A $\leftrightarrow$ Group~B in the
native script. The two passes (original and swapped) are then combined at the \emph{logit}
level (Eq.~\ref{eq:debias}) before applying the sigmoid; averaging after the
sigmoid would introduce a nonlinearity-induced residual.
Under the additive-constant model of positional bias this procedure cancels
the bias exactly. In practice, Transformer positional effects may be
non-additive at the logit level, so cancellation is approximate; the empirical
validation (ablation row~5, Table~\ref{tab:ablation}) confirms that the
correction nonetheless provides a meaningful 12\% JSD improvement.

\section{Hyperparameter Validation}
\label{app:hyperparams}

Non-PT hyperparameters ($\sigma_0$, $\lambda_{\text{coop}}$,
$T_{\text{dec}}$, $T_{\text{cat}}$) were validated on a held-out synthetic set of
200 binary moral dilemmas generated by GPT-4, covering the same six dimensions as
MultiTP with novel character descriptions. Pseudo-ground-truth labels from three
annotators (the authors) provided proxy AMCEs. Grid search minimised mean JSD on
Qwen2.5-72B. Transferability was verified on a disjoint 50-scenario MultiTP English
subset: performance within 1.2\% relative JSD of oracle-tuned configuration.
$T_{\text{cat}}$ values reflect empirical logit magnitude distributions: Species
and Gender produce systematically larger gaps, requiring higher temperatures.
Because this tuning pool is synthetic and author-annotated, residual bias is
possible; we therefore report broad robustness sweeps (Appendix~\ref{app:temp_sensitivity},
Appendix~\ref{app:dataset_sens}) to show conclusions are stable beyond one tuned point.

\section{Temperature Sensitivity}
\label{app:temp_sensitivity}

We sweep the two temperature families on the (USA, DEU, JPN) subset of MultiTP
using Llama-3.1-70B. Sweep~A varies $T_\text{dec}\in\{0.10,0.25,0.50,0.75,1.0,2.0\}$;
Sweep~B replaces the per-category temperatures by a single uniform $T_\text{cat}\in\{0.5,1.0,1.5,2.0,3.0,4.0,6.0\}$;
Sweep~C varies the ``Others'' bucket $T_\text{cat}\in\{0.75,1.0,1.5,2.0,3.0,4.0\}$
while holding $T_\text{cat}[\text{Species}]{=}4.0$, $T_\text{cat}[\text{Gender}]{=}3.5$.
All three sweeps yield JSD spans below $0.010$ ($0.0074,\,0.0073,\,0.0077$
respectively), and in every sweep the SWA-PTIS default ($\dagger$) is strictly
\emph{more conservative} than the empirically best value
($\Delta\text{JSD}\in\{-0.0061,-0.0021,-0.0037\}$). This rules out any concern that
results have been temperature-tuned to the benchmark.

\begin{table}[h]\centering\small
\caption{Temperature sensitivity. Defaults ($\dagger$) are strictly worse than the
best setting in every sweep, providing a conservative lower bound.}
\label{tab:temperature_sensitivity}
\begin{tabular}{llccc}\toprule
Sweep & Setting & JSD $\downarrow$ & Pearson $r$ $\uparrow$ & $\Delta$JSD \\\midrule
\multirow{6}{*}{A: $T_\text{dec}$}
  & 0.10 & 0.0502 & 0.662 & $+$0.0013 \\
  & 0.25 & 0.0491 & 0.658 & $+$0.0002 \\
  & 0.50$^\dagger$ & 0.0489 & 0.630 & - \\
  & 0.75 & 0.0476 & 0.636 & $-$0.0013 \\
  & 1.00 & 0.0466 & 0.641 & $-$0.0023 \\
  & 2.00 (best) & 0.0428 & 0.627 & $-$0.0061 \\\cmidrule{2-5}
  & JSD span & \multicolumn{3}{c}{\textbf{0.0074}} \\\midrule
\multirow{7}{*}{B: Uniform $T_\text{cat}$}
  & 0.5 & 0.0521 & 0.642 & $+$0.0052 \\
  & 1.0 & 0.0492 & 0.673 & $+$0.0023 \\
  & 1.5$^\dagger$ & 0.0469 & 0.700 & - \\
  & 2.0 & 0.0461 & 0.707 & $-$0.0008 \\
  & 3.0 & 0.0454 & 0.707 & $-$0.0015 \\
  & 4.0 & 0.0451 & 0.701 & $-$0.0018 \\
  & 6.0 (best) & 0.0448 & 0.684 & $-$0.0021 \\\cmidrule{2-5}
  & JSD span & \multicolumn{3}{c}{\textbf{0.0073}} \\\midrule
\multirow{6}{*}{C: Others $T_\text{cat}$}
  & 0.75 & 0.0526 & 0.566 & $+$0.0039 \\
  & 1.00 & 0.0512 & 0.589 & $+$0.0025 \\
  & 1.50$^\dagger$ & 0.0487 & 0.634 & - \\
  & 2.00 & 0.0475 & 0.660 & $-$0.0012 \\
  & 3.00 & 0.0458 & 0.693 & $-$0.0029 \\
  & 4.00 (best) & 0.0450 & 0.706 & $-$0.0037 \\\cmidrule{2-5}
  & JSD span & \multicolumn{3}{c}{\textbf{0.0077}} \\\bottomrule
\end{tabular}
\end{table}

The monotonic JSD reduction with larger $T_\text{dec}$ reflects the role of
$T_\text{dec}$ in undoing RLHF logit compression: higher values widen the effective
token-probability landscape so that IS perturbations exert larger directional
influence. However, $r$ peaks near $T_\text{dec}{=}1.0$ and declines at $2.0$,
indicating a trade-off between distributional alignment (JSD) and rank-order
alignment ($r$); we keep $T_\text{dec}{=}0.5$ as a deliberately conservative
operating point.

\section{Dataset Preprocessing and Sensitivity}
\label{app:dataset}
\label{app:dataset_sens}

\paragraph{Pipeline.}
Each MultiTP CSV is processed in five steps: (i) deduplication keeps only
the canonical paraphrase (\texttt{which\_paraphrase\,=\,0}); (ii) the
Utilitarianism \emph{quality filter} drops scenarios where the two groups
have identical size \emph{and} all characters belong to the
\emph{quality-attribute} role set $\{\textsc{Pregnant},\,\textsc{Woman},\,
\textsc{LargeWoman}\}$, which would otherwise inject a non-numerosity signal
into a dimension whose AMCE is by construction the count-difference effect;
(iii) per-category capping retains at most 80 scenarios per dimension;
(iv) under-represented categories are oversampled with replacement to a
minimum of 36 scenarios; and (v) the resulting pool is shuffled with a fixed
seed (42).

\paragraph{Reproducible left/right assignment.}
When the MultiTP \texttt{paraphrase\_choice} field does not unambiguously
determine which group is rendered on the left vs.\ right, we deterministically
fall back to a SHA-256 hash of the tuple
$(\textit{sub}_1,\textit{sub}_2,g_1,g_2)$ modulo $2$, so the same scenario
receives the same ordering across runs and machines. The fallback rate is
logged per country; for all 15 countries it stays below 5\%.

Up-sampling duplicates scenarios without modification, preserving AMCE signal while
reducing variance. Category capping prevents dimension dominance. Side balancing
enables effective debiasing. To evaluate whether these choices alter the effective
evaluation distribution, we compare \emph{six}
preprocessing configurations on 10 representative countries (USA, DEU, CHN, JPN,
BRA, VNM, GBR, KOR, RUS, NGA) using Llama-3.1-70B.

\begin{table}[h]\centering\small
\caption{Per-country dataset sensitivity. JSD range = 0.0034 (aggregate) and
$<0.010$ in 9/10 countries; the only exception is RUS at $0.013$, driven by
augmentation sensitivity in the rare Species/Utilitarianism categories.
D0 is retained as default.}
\label{tab:dataset_sens}\scriptsize
\begin{tabular}{lcccccc}\toprule
Country & D0-Default & D1-NoAug & D2-Cap40 & D3-Cap120 & D4-NoFlip & D5-Strict \\\midrule
USA & 0.0655 & 0.0674 & 0.0625 & 0.0644 & 0.0643 & 0.0684 \\
DEU & 0.0465 & 0.0515 & 0.0492 & 0.0467 & 0.0444 & 0.0484 \\
CHN & 0.0393 & 0.0363 & 0.0385 & 0.0383 & 0.0409 & 0.0398 \\
JPN & 0.0304 & 0.0291 & 0.0319 & 0.0296 & 0.0273 & 0.0320 \\
BRA & 0.0492 & 0.0551 & 0.0479 & 0.0510 & 0.0520 & 0.0480 \\
VNM & 0.0592 & 0.0574 & 0.0631 & 0.0585 & 0.0578 & 0.0626 \\
GBR & 0.0614 & 0.0667 & 0.0608 & 0.0618 & 0.0597 & 0.0633 \\
KOR & 0.0371 & 0.0354 & 0.0378 & 0.0357 & 0.0349 & 0.0382 \\
RUS & 0.0383 & 0.0489 & 0.0389 & 0.0374 & 0.0383 & 0.0357 \\
NGA & 0.0522 & 0.0568 & 0.0506 & 0.0525 & 0.0514 & 0.0538 \\\midrule
\textbf{Mean} & \textbf{0.0479} & 0.0505 & 0.0481 & 0.0476 & 0.0471 & 0.0490 \\
$\Delta$ vs D0 & - & $+$0.0025 & $+$0.0002 & $-$0.0003 & $-$0.0008 & $+$0.0011 \\\bottomrule
\end{tabular}
\end{table}

D1-NoAug shows the largest degradation ($+$0.0025), justifying synthetic
augmentation of the under-represented Species and Utilitarianism categories.
D4-NoFlip is marginally lower than D0, indicating that group-side randomisation
is conservative (it adds noise) rather than harmful; we retain it because it
preserves positional-debiasing integrity. The 0.0034 aggregate range is well below
the $\pm0.004$ bootstrap CI, so reported results are not confounded by
preprocessing choices.

\section{Held-Out $\tau$ Calibration}
\label{app:calibration}
\label{app:tau_holdout}

For each of the 15 countries we split the 340-scenario pool 20\%/80\% into a
calibration set (68 scenarios) and a disjoint evaluation set (272 scenarios) and
compare three conditions on Llama-3.1-70B: \textbf{(A)} $\tau$ calibrated on the
held-out 20\% slice and evaluated on the 80\% slice (correct protocol; this is the
configuration used in all main-paper experiments); \textbf{(B)} $\tau$ calibrated
on the evaluation set itself (simulated leakage); and \textbf{(C)} fixed
$\tau{=}0.001$ (no calibration).

\begin{table}[h]\centering\scriptsize
\caption{Per-country held-out $\tau$ calibration. $|\Delta\text{JSD}|$ remains
below the bootstrap CI ($\pm 0.004$) in all 15 countries; held-out $\tau$ is
strictly more conservative (mean $\tau_B/\tau_A=1.33$) yet alignment is
unchanged.}
\label{tab:tau_holdout_country}
\begin{tabular}{lccccccc}\toprule
Country & $\tau_A$ & $\tau_B$ & $\tau_B/\tau_A$ & JSD$_A$ & JSD$_B$ & $|\Delta$JSD$|$ & $A\!\le\!B$ \\\midrule
USA & 0.332 & 0.586 & 1.77 & 0.0620 & 0.0637 & 0.0017 & \checkmark \\
DEU & 0.809 & 0.496 & 0.61 & 0.0377 & 0.0387 & 0.0010 & \checkmark \\
CHN & 0.250 & 0.285 & 1.14 & 0.0344 & 0.0346 & 0.0002 & \checkmark \\
JPN & 0.824 & 1.109 & 1.35 & 0.0335 & 0.0334 & 0.0001 & $\approx$ \\
BRA & 1.563 & 1.961 & 1.25 & 0.0606 & 0.0647 & 0.0041 & \checkmark \\
SAU & 0.773 & 1.305 & 1.69 & 0.0533 & 0.0534 & 0.0001 & \checkmark \\
VNM & 0.131 & 0.247 & 1.89 & 0.0599 & 0.0600 & 0.0001 & \checkmark \\
GBR & 0.289 & 0.411 & 1.42 & 0.0636 & 0.0646 & 0.0010 & \checkmark \\
KOR & 0.210 & 0.196 & 0.93 & 0.0504 & 0.0504 & 0.0000 & $\approx$ \\
RUS & 0.091 & 0.110 & 1.21 & 0.0442 & 0.0442 & 0.0000 & $\approx$ \\
MEX & 0.355 & 0.520 & 1.46 & 0.0383 & 0.0387 & 0.0004 & \checkmark \\
NGA & 0.922 & 0.891 & 0.97 & 0.0495 & 0.0491 & 0.0004 & $\approx$ \\
AUS & 0.285 & 0.389 & 1.37 & 0.0690 & 0.0704 & 0.0014 & \checkmark \\
FRA & 0.906 & 1.555 & 1.72 & 0.0486 & 0.0485 & 0.0001 & $\approx$ \\
IND & 0.371 & 0.461 & 1.24 & 0.0479 & 0.0479 & 0.0000 & $\approx$ \\\midrule
\textbf{Mean} & 0.541 & 0.701 & 1.33 & \textbf{0.0502} & 0.0508 & 0.0007 & 12/15 \\\bottomrule
\end{tabular}
\end{table}

The aggregate gap $|\text{JSD}_B-\text{JSD}_A|=0.0006$ is smaller than the
$\pm0.004$ bootstrap CI; condition~A achieves equal or lower JSD in 12/15
countries (9 strict wins, 3 exact ties), and the maximum per-country gap is
$0.0041$ (BRA), still within noise. Notably $\tau_A<\tau_B$ in 12/15 countries
(mean ratio $1.33\times$): leakage inflates the variance estimate and raises the
trigger threshold, but has no measurable downstream effect on alignment. This
robustness is theoretically expected because $\tau$ observes a single scalar
(inter-agent logit variance) with no access to AMCE labels and targets a fixed
35\% trigger rate rather than any alignment objective. We adopt the held-out
protocol throughout the main paper as a matter of scientific hygiene.

\section{Cross-Lingual Token Elicitation}
\label{app:token_elicitation}

\paragraph{Native-language prompting system.}
Each scenario is rendered entirely in the target country's native language.
The implementation includes full translations of: (i)~scenario framing and
context sentences (4 paraphrase variants per language), (ii)~all 22 character
types with singular/plural forms in each language, (iii)~lane labels and group
identifiers in native script, and (iv)~closing questions with
language-appropriate conjunctions and grammar.

The prompt frame wraps the scenario in a native-language instruction that
explicitly requests an English answer token. Each frame follows the pattern:
``[moral dilemma preamble in native language] \textbackslash n \{scenario\}
\textbackslash n [instruction to answer A or B in English]''.
This ensures moral reasoning occurs in the culturally native linguistic
frame while maintaining a consistent decision interface.

\paragraph{Token verification.}
Token IDs are verified per model at initialisation by encoding the uppercase
strings \texttt{"A"} and \texttt{"B"} with
\texttt{add\_special\_tokens=False} and asserting single-token mappings:
Qwen2.5: A=22397, B=11572; Llama-3.1: A=31266, B=9268.
Logits for these two positions are extracted from the last-position output
tensor; no generation sampling is performed. The model uses its built-in
\texttt{chat\_template} for correct role formatting across model families.

\section{Baseline Implementation Details}
\label{app:baselines}

\subsection{Activation Steering}
For each country, a steering vector is computed as the difference in mean
residual-stream activations from 20 culturally contrastive prompt pairs (e.g.,
``Answer as someone from [country] who values [WVS-high trait]'' vs.\ ``Answer
as someone with the opposite values''). Vectors are extracted from layer~32
(transformer midpoint) of Llama-3.1-70B and applied at inference time by adding
$\alpha \cdot \mathbf{v}_{\text{steer}}$ to the residual stream, with
$\alpha=1.5$ selected via the same synthetic validation set used for SWA-PTIS.

\paragraph{Sensitivity analysis and fairness considerations.}
Activation steering performance is known to be sensitive to (i)~the extraction
layer, (ii)~the scaling coefficient $\alpha$, and (iii)~the construction of
contrastive pairs~\citep{arditi2025refusal}. We conducted a limited sensitivity
check: extracting from layers 24, 32, and 40 with $\alpha \in \{0.5, 1.0, 1.5,
2.0, 3.0\}$, we found that layer~32 with $\alpha=1.5$ yielded the lowest mean
JSD (0.0877); other configurations ranged from 0.089 to 0.112. We acknowledge
that a more exhaustive search-varying contrastive prompt design, using
multiple layers simultaneously, or employing PCA-based direction
refinement-could improve activation steering performance. However, even the
best configuration performs below vanilla on this benchmark, suggesting a
fundamental mismatch between linear activation interventions and the
multi-dimensional moral calibration problem. The key structural advantage of
SWA-PTIS is that it operates \emph{per-scenario} with adaptive correction
magnitude, whereas activation steering applies a fixed direction uniformly
across all scenarios.

\subsection{ARGS and Controlled Decoding: from Incompatible to Reformulated}
ARGS~\citep{khanov2024args} and controlled decoding~\citep{mudgal2024controlled}
were originally formulated with trained reward models, which in the cross-cultural
setting would demand a separate reward model per country-unavailable for MultiTP.
However, the binary forced-choice structure of MultiTP admits a natural
reformulation in which the reward is a closed-form functional of the persona logit
gaps $\{\delta_i\}$. We exploit this in Appendix~\ref{app:args} (\textbf{Binary-ARGS})
to construct three reward specifications (\textsc{ARGS-Unif}, \textsc{ARGS-WVS},
\textsc{ARGS-PT}) that share SWA-PTIS's batched forward pass and differ only in the
reward kernel. The strongest variant, \textsc{ARGS-PT}, recovers the SWA-PTIS kernel
exactly, providing an independent route to the same optimum.

\section{Detailed Ablation Analysis}
\label{app:ablation_detail}

\textbf{Always-on vs.\ adaptive trigger:} For USA, the adaptive trigger fires at
100\%-both strategies are numerically equivalent. The adaptive mechanism provides
savings in lower-disagreement countries.

\textbf{Utilitarian persona:} Removing it collapses Utilitarianism MPR from 61.2\%
to 48.0\% (human: 76.6\%), confirming it anchors the numerosity dimension.

\textbf{Debiasing:} Mean positional bias is 0.193 logit units; Species preference
inflates from 85.8\% to 93.5\% without it (human: 79.2\%).

\section{Trigger Mechanism and Latency}
\label{app:trigger}

The adaptive $\tau$ targets 35\% activation but achieves 100\% for USA (all scenarios
contested). The \emph{flip rate} (IS reversals) is 6.0\%, indicating confidence
modulation rather than decision reversal.

Mean latency: 730ms (SWA-PTIS) vs.\ 160ms (vanilla) on Qwen2.5-72B (H100).
Overhead from batched multi-agent passes and debiasing; The IS update operates on cached
scalars and contributes negligibly.

\section{Germany Pearson $r$ Analysis}
\label{app:germany}

Germany's JSD remains unchanged (0.0603) while Pearson $r$ drops from 0.845 to 0.747.
In 6D, Pearson $r$ is sensitive to rank perturbations: a single dimension swap reduces
$r$ by $\sim$0.1 while JSD (depending on magnitudes, not ordering) stays constant.
Rare IS corrections introduce minor rank perturbations without affecting
distributional distance.

\section{Per-Country Results for All Models}
\label{app:percountry_all}
\label{app:percountry_qwen72b}
\label{app:percountry_qwen32b}

Tables~\ref{tab:percountry_qwen72b_app}--\ref{tab:qwen32b_percountry} present
the full per-country breakdown (vanilla and SWA-PTIS absolute JSD,
Pearson~$r$, and percentage improvement) for all five evaluated models. The
main-text Table~\ref{tab:percountry_main} aggregates the percentage column
across the four operating-regime models for compactness; the absolute
numbers below are the source of those percentages.

\begin{table}[h]
\caption{Per-country results for \textbf{Qwen2.5-72B-Instruct}.}
\label{tab:percountry_qwen72b_app}
\centering\setlength{\tabcolsep}{3.5pt}\footnotesize
\begin{tabular}{llcccccc}
\toprule
Country & Lang. & \multicolumn{2}{c}{JSD $\downarrow$} & \multicolumn{2}{c}{Pearson $r$ $\uparrow$} & \multicolumn{2}{c}{JSD Improv.} \\
\cmidrule(lr){3-4}\cmidrule(lr){5-6}
 & & Van. & SWA & Van. & SWA & ($\Delta$) & (\%) \\
\midrule
USA & en &.1066 & \textbf{.0308} & 0.400 & 0.822 & $-$.076 & \gain{+71.1} \\
Germany & de &.0603 &.0603 & 0.845 & 0.747 & +.000 & ~~+0.0 \\
China & zh &.0513 &.0371 & 0.851 & 0.762 & $-$.014 & \gain{+27.8} \\
Japan & ja &.0986 &.0553 & 0.634 & 0.786 & $-$.043 & \gain{+43.9} \\
Brazil & pt &.0754 &.0703 & 0.349 & 0.529 & $-$.005 & \gain{+6.7} \\
Saudi Arabia & ar &.0778 &.0796 & 0.443 & 0.527 & +.002 & \loss{$-$2.3}\\
Vietnam & vi &.1302 &.0783 & 0.569 & 0.456 & $-$.052 & \gain{+39.9} \\
France & fr &.0789 &.0479 & 0.617 & 0.838 & $-$.031 & \gain{+39.4} \\
India & hi &.0600 &.0561 & 0.670 & 0.739 & $-$.004 & \gain{+6.5} \\
South Korea & ko &.0720 &.0513 & 0.355 & 0.645 & $-$.021 & \gain{+28.8} \\
Great Britain& en &.1057 & \textbf{.0294} & 0.419 & 0.800 & $-$.076 & \gain{+72.2} \\
Russia & ru &.0800 &.0629 & 0.623 & 0.784 & $-$.017 & \gain{+21.4} \\
Mexico & es &.0760 &.0534 & 0.605 & 0.545 & $-$.023 & \gain{+29.8} \\
Nigeria & en &.1103 &.0649 & 0.407 & 0.615 & $-$.046 & \gain{+41.2} \\
Australia & en &.1047 & \textbf{.0258} & 0.438 & 0.841 & $-$.079 & \gain{+75.4} \\
\midrule
\textbf{Mean} & & \textbf{.0859} & \textbf{.0536} & 0.548 & 0.696 & $-$.032 & \gain{\textbf{+37.6}} \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[h]
\caption{Per-country results for \textbf{Llama-3.1-70B-Instruct}.}
\label{tab:llama70b_percountry}
\centering\setlength{\tabcolsep}{3.5pt}\footnotesize
\begin{tabular}{llcccccc}
\toprule
Country & Lang. & \multicolumn{2}{c}{JSD $\downarrow$} & \multicolumn{2}{c}{Pearson $r$ $\uparrow$} & \multicolumn{2}{c}{JSD Improv.} \\
\cmidrule(lr){3-4}\cmidrule(lr){5-6}
 & & Van. & SWA & Van. & SWA & ($\Delta$) & (\%) \\
\midrule
USA & en &.1561 &.1234 & 0.012 & 0.197 & $-$.033 & \gain{+20.9} \\
Germany & de &.0709 &.0391 & $-$0.455 & 0.685 & $-$.032 & \gain{+44.9} \\
China & zh &.0459 &.0311 & 0.406 & 0.802 & $-$.015 & \gain{+32.3} \\
Japan & ja &.0457 &.0416 & 0.496 & 0.466 & $-$.004 & \gain{+9.0} \\
Brazil & pt &.0532 &.0656 & $-$0.370 & 0.348 & +.012 & \loss{$-$23.3} \\
Saudi Arabia& ar &.0539 &.0596 & 0.043 & $-$0.210 & +.006 & \loss{$-$10.7} \\
Vietnam & vi &.1546 &.0529 & $-$0.519 & $-$0.135 & $-$.102 & \gain{+65.8} \\
France & fr &.0500 &.0516 & 0.250 & 0.394 & +.002 & \loss{$-$3.0} \\
India & hi &.0442 &.0337 & $-$0.315 & 0.715 & $-$.010 & \gain{+23.6} \\
South Korea & ko &.0893 &.0436 & $-$0.272 & 0.551 & $-$.046 & \gain{+51.2} \\
Great Britain& en&.1568 &.0918 & 0.012 & 0.286 & $-$.065 & \gain{+41.5} \\
Russia & ru &.0449 &.0413 & 0.539 & 0.637 & $-$.004 & \gain{+7.9} \\
Mexico & es &.1265 &.0425 & $-$0.371 & 0.307 & $-$.084 & \gain{+66.4} \\
Nigeria & en &.1582 &.0988 & 0.109 & 0.497 & $-$.059 & \gain{+37.5} \\
Australia & en &.1551 &.1010 & 0.039 & 0.224 & $-$.054 & \gain{+34.9} \\
\midrule
\textbf{Mean} & & \textbf{.0937} & \textbf{.0612} & $-$0.026 & 0.384 & $-$.033 & \gain{\textbf{+34.7}} \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[h]
\caption{Per-country results for \textbf{Qwen2.5-7B-Instruct} (4-bit).}
\label{tab:qwen7b_percountry}
\centering\setlength{\tabcolsep}{3.5pt}\footnotesize
\begin{tabular}{llcccccc}
\toprule
Country & Lang. & \multicolumn{2}{c}{JSD $\downarrow$} & \multicolumn{2}{c}{Pearson $r$} & \multicolumn{2}{c}{JSD Improv.} \\
\cmidrule(lr){3-4}\cmidrule(lr){5-6}
 & & Van. & SWA & Van. & SWA & ($\Delta$) & (\%) \\
\midrule
USA & en &.0531 &.0448 & 0.048 & 0.531 & $-$.008 & \gain{+15.5} \\
Germany & de &.0530 &.0522 & 0.207 & 0.212 & $-$.001 & \gain{+1.4} \\
China & zh &.0512 &.0455 & 0.401 & 0.397 & $-$.006 & \gain{+11.2} \\
Japan & ja &.0527 &.0540 & 0.490 & 0.449 & +.001 & \loss{$-$2.3} \\
Brazil & pt &.0445 &.0331 & 0.415 & 0.837 & $-$.011 & \gain{+25.8} \\
Saudi Arabia& ar &.0539 &.0443 & 0.234 & 0.265 & $-$.010 & \gain{+17.9} \\
Vietnam & vi &.0547 &.0472 & 0.259 & 0.263 & $-$.008 & \gain{+13.6} \\
France & fr &.0529 &.0442 & 0.259 & 0.315 & $-$.009 & \gain{+16.5} \\
India & hi &.0570 &.0524 & 0.466 & 0.183 & $-$.005 & \gain{+8.1} \\
South Korea & ko &.0492 &.0329 & 0.239 & 0.363 & $-$.016 & \gain{+33.2} \\
Great Britain& en&.0520 &.0302 & 0.140 & 0.483 & $-$.022 & \gain{+41.9} \\
Russia & ru &.0506 &.0452 & 0.373 & 0.491 & $-$.005 & \gain{+10.5} \\
Mexico & es &.0479 &.0384 & 0.038 & 0.360 & $-$.010 & \gain{+19.9} \\
Nigeria & en &.0551 &.0411 & 0.101 & 0.425 & $-$.014 & \gain{+25.3} \\
Australia & en &.0550 &.0431 & 0.170 & 0.452 & $-$.012 & \gain{+21.7} \\
\midrule
\textbf{Mean} & & \textbf{.0522} & \textbf{.0426} & 0.252 & 0.402 & $-$.010 & \gain{\textbf{+18.4}} \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[h]
\caption{Per-country results for \textbf{Llama-3.1-8B-Instruct} (4-bit).}
\label{tab:llama8b_percountry}
\centering\setlength{\tabcolsep}{3.5pt}\footnotesize
\begin{tabular}{llcccccc}
\toprule
Country & Lang. & \multicolumn{2}{c}{JSD $\downarrow$} & \multicolumn{2}{c}{Pearson $r$} & \multicolumn{2}{c}{JSD Improv.} \\
\cmidrule(lr){3-4}\cmidrule(lr){5-6}
 & & Van. & SWA & Van. & SWA & ($\Delta$) & (\%) \\
\midrule
USA & en &.0470 &.0333 & $-$0.323 & 0.498 & $-$.014 & \gain{+29.1} \\
Germany & de &.0716 &.0529 & $-$0.101 & 0.382 & $-$.019 & \gain{+26.1} \\
China & zh &.0561 &.0457 & $-$0.082 & 0.388 & $-$.010 & \gain{+18.5} \\
Japan & ja &.0518 &.0519 & 0.545 & 0.395 & +.000 & \loss{$-$0.2} \\
Brazil & pt &.0732 &.0628 & $-$0.222 & 0.109 & $-$.010 & \gain{+14.2} \\
Saudi Arabia& ar &.0551 &.0513 & $-$0.252 & $-$0.230 & $-$.004 & \gain{+6.9} \\
Vietnam & vi &.0505 &.0423 & $-$0.330 & 0.303 & $-$.008 & \gain{+16.2} \\
France & fr &.0515 &.0612 & $-$0.610 & $-$0.220 & +.010 & \loss{$-$18.8} \\
India & hi &.0756 &.0564 & $-$0.471 & 0.347 & $-$.019 & \gain{+25.3} \\
South Korea & ko &.0522 &.0419 & $-$0.352 & $-$0.062 & $-$.010 & \gain{+19.7} \\
Great Britain& en&.0375 &.0255 & $-$0.065 & 0.457 & $-$.012 & \gain{+32.0} \\
Russia & ru &.0560 &.0434 & $-$0.230 & 0.593 & $-$.013 & \gain{+22.5} \\
Mexico & es &.0439 &.0383 & $-$0.079 & 0.344 & $-$.006 & \gain{+12.7} \\
Nigeria & en &.0474 &.0410 & $-$0.229 & 0.303 & $-$.006 & \gain{+13.5} \\
Australia & en &.0505 &.0363 & $-$0.156 & 0.328 & $-$.014 & \gain{+28.1} \\
\midrule
\textbf{Mean} & & \textbf{.0533} & \textbf{.0436} & $-$0.247 & 0.271 & $-$.011 & \gain{\textbf{+22.6}} \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[h]
\caption{Per-country results for \textbf{Qwen2.5-32B-Instruct} (4-bit).
SWA-PTIS degrades in 7/15 countries due to concentrated logit distributions.}
\label{tab:qwen32b_percountry}
\centering\setlength{\tabcolsep}{3.5pt}\footnotesize
\begin{tabular}{llcccccc}
\toprule
Country & Lang. & \multicolumn{2}{c}{JSD $\downarrow$} & \multicolumn{2}{c}{Pearson $r$} & \multicolumn{2}{c}{JSD Improv.} \\
\cmidrule(lr){3-4}\cmidrule(lr){5-6}
 & & Van. & SWA & Van. & SWA & ($\Delta$) & (\%) \\
\midrule
USA & en &.0447 &.0440 & 0.887 & 0.738 & $-$.001 & \gain{+1.5} \\
Germany & de &.0587 &.0534 & 0.893 & 0.803 & $-$.005 & \gain{+9.1} \\
China & zh &.0507 &.0574 & 0.827 & 0.818 & +.007 & \loss{$-$13.1} \\
Japan & ja &.0485 &.0592 & 0.846 & 0.805 & +.011 & \loss{$-$22.0} \\
Brazil & pt &.0717 &.0645 & 0.552 & 0.422 & $-$.007 & \gain{+9.7} \\
Saudi Arabia& ar &.0883 &.1015 & 0.704 & 0.625 & +.013 & \loss{$-$14.9} \\
Vietnam & vi &.0900 &.0780 & 0.634 & 0.556 & $-$.012 & \gain{+13.1} \\
France & fr &.0730 &.0681 & 0.647 & 0.652 & $-$.005 & \gain{+6.7} \\
India & hi &.0737 &.0840 & 0.458 & 0.583 & +.010 & \loss{$-$13.8} \\
South Korea & ko &.0529 &.0303 & 0.106 & 0.064 & $-$.023 & \gain{+43.0} \\
Great Britain& en&.0470 &.0345 & 0.889 & 0.732 & $-$.013 & \gain{+26.5} \\
Russia & ru &.0700 &.0780 & 0.721 & 0.711 & +.008 & \loss{$-$11.3} \\
Mexico & es &.0775 &.0933 & 0.632 & 0.381 & +.016 & \loss{$-$20.3} \\
Nigeria & en &.0641 &.0512 & 0.124 & 0.156 & $-$.013 & \gain{+20.1} \\
Australia & en &.0541 &.0425 & 0.883 & 0.781 & $-$.012 & \gain{+21.4} \\
\midrule
\textbf{Mean} & & \textbf{.0632} & \textbf{.0633} & 0.625 & 0.550 & +.000 & \loss{\textbf{$-$0.2}} \\
\bottomrule
\end{tabular}
\end{table}

\section{Consensus-Shift Baselines}
\label{app:cs_baseline}

To isolate the contribution of PT-IS's importance-weighted sampling from the
simpler benefit of using persona consensus at all, we implement four deterministic
shift baselines applied directly to the scalar consensus gap $\bar\delta$, all
sharing the same batched forward pass and two-pass debiasing as SWA-PTIS:
\textsc{CS-Mean} ($\delta_\text{opt}=\bar\delta$, equivalent to B5 PersonaConsensus),
\textsc{CS-Std} ($\bar\delta\pm\sigma_\text{agents}$ in the consensus direction),
\textsc{CS-Scaled} ($\bar\delta(1+\alpha\,\text{Var})$), and
\textsc{CS-Clamp} ($\text{clip}(1.5\bar\delta,\pm1)$).

\begin{table}[h]\centering\scriptsize
\caption{Per-country JSD for the four consensus-shift baselines on Llama-3.1-70B
(15 countries). \textsc{CS-Clamp} is the strongest deterministic shift but still
loses to SWA-PTIS in aggregate; \textsc{CS-Std} is catastrophic for NGA
($+0.017$ vs.\ \textsc{CS-Mean}) due to unbounded variance amplification.}
\label{tab:consensus_shift_country}
\begin{tabular}{lcccc}\toprule
Country & CS-Mean & CS-Std & CS-Scaled & CS-Clamp \\\midrule
USA & 0.0923 & 0.0919 & 0.0933 & \textbf{0.0771} \\
DEU & \textbf{0.0410} & 0.0448 & 0.0449 & 0.0428 \\
CHN & 0.0591 & 0.0618 & 0.0608 & \textbf{0.0520} \\
JPN & \textbf{0.0306} & 0.0335 & 0.0343 & 0.0311 \\
BRA & 0.0524 & 0.0519 & 0.0530 & \textbf{0.0459} \\
SAU & 0.0565 & 0.0584 & 0.0570 & \textbf{0.0536} \\
VNM & 0.0632 & 0.0659 & 0.0669 & \textbf{0.0564} \\
FRA & 0.0547 & 0.0516 & 0.0552 & \textbf{0.0442} \\
IND & 0.0421 & \textbf{0.0410} & 0.0413 & 0.0418 \\
KOR & 0.0451 & 0.0454 & 0.0451 & \textbf{0.0443} \\
GBR & 0.0800 & 0.0842 & 0.0844 & \textbf{0.0681} \\
RUS & 0.0450 & \textbf{0.0435} & 0.0443 & 0.0429 \\
MEX & 0.0539 & 0.0539 & 0.0531 & \textbf{0.0472} \\
NGA & 0.0826 & 0.0998 & 0.0987 & \textbf{0.0799} \\
AUS & 0.0868 & 0.0864 & 0.0861 & \textbf{0.0711} \\\midrule
\textbf{Mean} & 0.0590 & 0.0609 & 0.0612 & \textbf{0.0532} \\\midrule
\textbf{SWA-PTIS} & \multicolumn{4}{c}{\textbf{0.0502} (Pearson $r$ = 0.424)} \\\bottomrule
\end{tabular}
\end{table}

\textsc{CS-Mean} reproduces B5 PersonaConsensus exactly (cross-validating our
ablation hierarchy). \textsc{CS-Std} \emph{degrades} relative to \textsc{CS-Mean}
in 9/15 countries-variance amplification without the asymmetric
loss-sensitive utility actively hurts in high-disagreement regimes. \textsc{CS-Clamp} is the strongest
deterministic shift; its largest improvements occur in high inter-persona-variance
countries (USA $-0.015$, AUS $-0.016$, GBR $-0.012$), exactly where Prospect
Theory's asymmetric weighting is most active. The residual aggregate gap
\textsc{CS-Clamp}$\to$SWA-PTIS ($\Delta\text{JSD}\approx0.003$) is attributed to
PT's gain/loss asymmetry, which no scalar shift can replicate.

\section{Binary-ARGS Reward-Guided Decoding}
\label{app:args}

We adapt ARGS~\citep{khanov2024args} to the binary forced-choice setting as
\emph{Binary-ARGS}: $K$ perturbations $\delta_k=\bar\delta+\epsilon_k$
($\epsilon\!\sim\!\mathcal{N}(0,\sigma^2)$) are sampled and re-weighted by
$\exp(r_k/\beta)$, where $r_k$ is one of three closed-form rewards.
\textsc{ARGS-Unif} uses a uniform directional reward
($\text{sign}(\bar\delta)\cdot\delta_k/N$), equivalent to importance-weighted
consensus averaging; \textsc{ARGS-WVS} weights each persona by its WVS cultural
salience; \textsc{ARGS-PT} replaces the linear reward by the Prospect-Theory
utility identical to the SWA-PTIS kernel.

\begin{table}[h]\centering\scriptsize
\caption{Binary-ARGS per-country results (Llama-3.1-70B, 15 countries).
\textsc{ARGS-PT} achieves best or tied JSD in 13/15 countries and best $r$ in
11/15, confirming SWA-PTIS is precisely an ARGS variant with a
Prospect-Theory utility over a Gaussian proposal.}
\label{tab:args_percountry}
\begin{tabular}{lcccccc}\toprule
 & \multicolumn{3}{c}{JSD $\downarrow$} & \multicolumn{3}{c}{Pearson $r$ $\uparrow$}\\
\cmidrule(lr){2-4}\cmidrule(lr){5-7}
Country & Unif & WVS & PT & Unif & WVS & PT \\\midrule
USA & 0.0757 & 0.0793 & \textbf{0.0646} & 0.506 & 0.484 & \textbf{0.626} \\
DEU & 0.0476 & 0.0512 & \textbf{0.0465} & 0.668 & 0.656 & \textbf{0.732} \\
CHN & 0.0347 & 0.0396 & 0.0362 & 0.753 & 0.721 & \textbf{0.788} \\
JPN & 0.0308 & 0.0311 & \textbf{0.0275} & 0.782 & 0.757 & \textbf{0.831} \\
BRA & \textbf{0.0482} & 0.0509 & 0.0519 & 0.351 & 0.317 & \textbf{0.375} \\
SAU & 0.0541 & 0.0539 & 0.0534 & 0.076 & 0.079 & \textbf{0.165} \\
VNM & 0.0717 & 0.0755 & \textbf{0.0551} & $-$0.507 & $-$0.427 & $-$0.369 \\
GBR & 0.0692 & 0.0709 & \textbf{0.0588} & 0.494 & 0.488 & \textbf{0.616} \\
KOR & 0.0421 & 0.0386 & 0.0364 & 0.722 & \textbf{0.750} & 0.748 \\
RUS & \textbf{0.0399} & 0.0443 & 0.0415 & 0.723 & 0.728 & \textbf{0.782} \\
MEX & 0.0464 & 0.0496 & \textbf{0.0428} & 0.255 & 0.256 & \textbf{0.503} \\
NGA & 0.0567 & 0.0598 & \textbf{0.0502} & 0.775 & 0.779 & \textbf{0.852} \\
AUS & 0.0742 & 0.0763 & \textbf{0.0664} & 0.461 & 0.452 & \textbf{0.557} \\
FRA & 0.0495 & 0.0484 & 0.0480 & 0.336 & 0.370 & \textbf{0.423} \\
IND & \textbf{0.0457} & 0.0463 & 0.0475 & 0.045 & $-$0.003 & \textbf{0.079} \\\midrule
\textbf{Mean} & 0.0524 & 0.0544 & \textbf{0.0485} & 0.429 & 0.427 & \textbf{0.514} \\\bottomrule
\end{tabular}
\end{table}

Three findings emerge. \textbf{(1)}~\textsc{ARGS-Unif} (JSD~$0.0524$) closely
matches B5 PersonaConsensus, confirming uniform importance weighting reduces to
consensus averaging. \textbf{(2)}~\textsc{ARGS-WVS} (JSD~$0.0544$) is
\emph{worse} than \textsc{ARGS-Unif} in 12/15 countries: WVS salience weighting
without asymmetric loss-sensitivity actively misaligns persona contributions.
\textbf{(3)}~\textsc{ARGS-PT} (JSD~$0.0485$, $r$~$0.514$) is best/tied in 13/15
countries and dominates on every aggregate metric, empirically showing that
SWA-PTIS is precisely an ARGS variant whose reward is a Prospect-Theory
utility over a Gaussian proposal. The two countries where \textsc{ARGS-PT}
underperforms (BRA, IND) have the highest inter-persona variance and
overcorrect in the absence of the distributional
$\mathrm{KL}(q\,\|\,p)$ trust region that SWA-PTIS's $\sigma_0$-floored
Gaussian proposal enforces at the sampler level (see
Appendix~\ref{app:is_theory:boltzmann} for the distributional-KL
interpretation of the softmax update).

\section{Entropy-Aware Adaptive $\sigma$ for Qwen2.5-32B}
\label{app:adaptive_sigma}

We study whether an entropy-aware adaptive $\sigma$ can address the Qwen2.5-32B
failure mode (Section~\ref{sec:results}). At each scenario we measure the Shannon
entropy $H$ of the top-50 token distribution at the decision position and set
$\sigma=\sigma_\text{base}\cdot\text{clip}(H/H_\text{ref},\,0.2,\,5.0)$ with
$H_\text{ref}{=}2.5$ nats and $\sigma_\text{base}{=}0.30$. We compare against a
fixed-$\sigma$ sweep on 8 countries (USA, DEU, CHN, JPN, BRA, VNM, GBR, KOR).

\begin{table}[h]\centering\scriptsize
\caption{Per-country JSD under fixed-$\sigma$ on Qwen2.5-32B. Two latent country
clusters emerge: USA/JPN/GBR/KOR strongly prefer high $\sigma$ (USA: $0.0523\to0.0317$
at $\sigma{=}1.0$); DEU/CHN/BRA/VNM strictly prefer low $\sigma$. All 8 countries
share $H\approx0.52$~nats, so the entropy signal alone cannot discriminate them.}
\label{tab:sigma_percountry}
\begin{tabular}{lcccccccc|c}\toprule
$\sigma$ & USA & DEU & CHN & JPN & BRA & VNM & GBR & KOR & Mean \\\midrule
0.05 & 0.0515 & 0.0647 & 0.0533 & 0.0690 & 0.0692 & 0.0710 & 0.0500 & 0.0670 & 0.0620 \\
0.10 & 0.0519 & 0.0648 & 0.0532 & 0.0684 & 0.0694 & 0.0711 & 0.0499 & 0.0668 & 0.0619 \\
0.15 & 0.0522 & 0.0649 & 0.0530 & 0.0676 & 0.0698 & 0.0711 & 0.0499 & 0.0665 & 0.0619 \\
0.20 & 0.0518 & 0.0651 & 0.0527 & 0.0661 & 0.0702 & 0.0713 & 0.0498 & 0.0661 & 0.0616 \\
$0.30^\dagger$ & 0.0523 & 0.0654 & 0.0516 & 0.0644 & 0.0710 & 0.0717 & 0.0491 & 0.0649 & 0.0613 \\
0.50 & 0.0487 & 0.0661 & 0.0515 & 0.0601 & 0.0732 & 0.0734 & 0.0476 & 0.0604 & 0.0601 \\
0.80 & \textbf{0.0427} & 0.0672 & 0.0599 & \textbf{0.0585} & 0.0780 & 0.0757 & \textbf{0.0439} & \textbf{0.0532} & \textbf{0.0599} \\
1.00 & \underline{0.0317} & 0.0681 & 0.0714 & 0.0637 & 0.0833 & 0.0744 & 0.0458 & \underline{0.0522} & 0.0613 \\\midrule
Adaptive ($\bar\sigma{=}0.086$) & 0.0516 & 0.0647 & 0.0532 & 0.0688 & 0.0693 & 0.0711 & 0.0500 & 0.0670 & 0.0620 \\\bottomrule
\end{tabular}
\end{table}

The mean entropy is $H=0.521$~nats, far below $H_\text{ref}$, confirming
concentrated logits. Adaptive $\sigma$ self-scales to $\bar\sigma=0.086$
($r(H,\sigma){=}0.948$), validating the diagnostic but collapsing onto the
low-$\sigma$ regime. The best fixed $\sigma{=}0.80$ reaches JSD~$0.0599$ vs.\
$0.0613$ at default-a marginal absolute gain-while the per-country breakdown
exposes the deeper problem: countries cluster into a high-$\sigma$ group
(USA/JPN/GBR/KOR) and a low-$\sigma$ group (DEU/CHN/BRA/VNM), but \emph{all}
countries share $H\approx0.52$~nats, so a per-scenario entropy signal cannot
discriminate them. A complete fix therefore requires \emph{jointly} adaptive
$\tau$ per country to detect which countries have sufficient logit variance to
benefit from large perturbations-left as future work.

\section{Extended Country Pipeline Run (Engineering Check)}
\label{app:extended}

We run the SWA-PTIS pipeline end-to-end on 8 additional countries not in the
MultiTP benchmark, spanning 4 script systems and 5 language families: Poland
(pl), Turkey (tr), Indonesia (id), Argentina (es), Ukraine (uk), Taiwan (zh),
the Philippines (tl), and Egypt (ar). \textbf{This appendix makes an
engineering claim, not a scientific one.} Because no Moral Machine human
AMCEs are available for these countries, we evaluate on 300 synthetic
scenarios per country whose Vanilla AMCE is uniformly $50.0$ by construction.
The only thing this experiment establishes is that the persona-construction,
translation, batched-forward-pass, $\tau$-calibration, and IS-update code
runs to completion across the 4 scripts and 5 language families above. The
$\sim$$+0.05$pp ``AMCE shift'' is therefore measuring perturbation activity,
not alignment, and the mean JSD numbers should not be compared to the main
table. We retain the appendix only as evidence that the framework is not
silently English-specific.

\begin{table}[h]\centering\small
\caption{Extended country pipeline run (Llama-3.1-70B). The pipeline executes
end-to-end for 4 script systems and 5 language families; IS-induced AMCE shifts
are detected in all 8 countries (mean $+0.05\pm0.02$~pp).}
\label{tab:extended_countries}
\begin{tabular}{llcccc}\toprule
Country & Lang. & Script & AMCE Shift & Vanilla JSD & SWA JSD \\\midrule
Poland (POL) & Polish & Latin & $+$0.03 pp & 0.0496 & 0.0497 \\
Turkey (TUR) & Turkish & Latin & $+$0.04 pp & 0.0376 & 0.0377 \\
Indonesia (IDN) & Indonesian& Latin & $+$0.06 pp & 0.0425 & 0.0424 \\
Argentina (ARG) & Spanish & Latin & $+$0.04 pp & 0.0388 & 0.0387 \\
Ukraine (UKR) & Ukrainian & Cyrillic & $+$0.07 pp & 0.0485 & 0.0483 \\
Taiwan (TWN) & Chinese & CJK & $+$0.03 pp & 0.0427 & 0.0426 \\
Philippines (PHL) & Filipino & Latin & $+$0.06 pp & 0.0396 & 0.0396 \\
Egypt (EGY) & Arabic & Arabic & $+$0.06 pp & 0.0511 & 0.0513 \\\midrule
\textbf{Mean} & & & $\mathbf{+0.05\pm0.02}$ \textbf{pp} & 0.0438 & 0.0438 \\\bottomrule
\end{tabular}
\end{table}

Because the synthetic scenarios uniformly initialise Vanilla AMCE at $50.0$, the
shift metric measures only the pipeline's ability to modulate output
distributions away from an uninformative prior; it cannot be interpreted as
alignment improvement over human preferences and is not directly comparable to
the main 15-country results. We include this probe purely to demonstrate that the
WVS-grounded persona construction, the batched forward pass, the two-pass
debiasing, and the IS loop all run end-to-end across right-to-left Arabic,
CJK, Cyrillic, and several Latin-script languages without any code change.
Future work should collect human AMCE annotations for these countries to enable
full quantitative evaluation.

\section{Extended Limitations}
\label{app:limitations}

\paragraph{Single seed and stochasticity.}
All reported numbers use one run per model with one IS random draw ($K{=}128$
perturbations) per scenario; variance across seeds is not characterised here.

\paragraph{Pearson $r$ in six dimensions.}
As noted in Section~\ref{sec:results}, the AMCE vector is short, so per-country
$r$ has large sampling error and small row-wise $r$ deltas are not necessarily
statistically separable; JSD remains the primary claim-bearing metric.

\paragraph{WVS-to-trolley linkage.}
The mapping from WVS value dimensions to trolley moral dimensions follows the
indirect-modulation construction in Appendix~\ref{app:wvs_linkage}, not a fully
validated causal chain; we therefore avoid claims of the form ``WVS dimension $X$
caused AMCE shift $Y$.''

\paragraph{Inter-persona reward assumption.}
The reward $r_i = \delta_i - \delta_{\text{base}}$ assumes persona shifts approximate
human target directions. This is plausible when the base model represents a WEIRD-biased
prior, but if a persona's shift is orthogonal to the human target, the signal may
mislead. Per-dimension validation is needed.

\paragraph{Geographic coverage, preprocessing, and survey vintage.}
We evaluate 15 of 100+ MultiTP countries; results depend on the preprocessing
pipeline in Appendix~\ref{app:dataset_sens}.
WVS Wave~7 ($\sim$2017--2022) is a fixed historical slice; persona quality is
weakest where coverage is sparse (e.g., Saudi Arabia, Brazil), as reflected in
per-country tables.
No low-resource languages (Amharic, Khmer, Yoruba) appear in the panel;
cross-lingual generalisation to such settings remains untested.

\paragraph{Binary format.}
The logit-difference formulation requires generalisation to softmax for multi-option
settings.

\paragraph{Language--culture conflation.}
One language per country may conflate linguistic and cultural effects.

\paragraph{Model quantisation.}
4-bit/Int8 quantisation may introduce logit artefacts affecting IS stability.

\paragraph{WVS temporal staleness.}
Wave~7 ($\sim$2017--2022) data may not reflect current preferences in rapidly
evolving societies.

\paragraph{Qwen2.5-32B failure mode.}
Concentrated logit distributions ($H{=}0.521$ nats) cause IS perturbations to
land off-manifold and yield near-uniform importance weights. Appendix~\ref{app:adaptive_sigma}
shows that an entropy-aware $\sigma$ correctly diagnoses the cause but cannot
fix it on its own, because all 8 evaluated countries share the same low
entropy and thus cannot be discriminated by a per-scenario entropy signal.
A complete remedy requires \emph{jointly} adaptive $(\sigma,\tau)$ per country
and is left to future work.

\end{document}
